---
title: "04_fineRADstructure_p"
author: "Nicole Pittoors"
date: "2026-08-25"
output: html_document
---

This script covers two related analyses:

1. **fineRADstructure co-ancestry** (Malinsky et al. 2018) for GoM10 (158 ind,
   9 pop) and Sub (174 ind, 10 pop). Two figure types are
   produced per dataset: a population-averaged co-ancestry matrix ordered
   west-to-east by geography, and an individual-level co-ancestry heatmap
   ordered by the fineRADstructure hierarchical clustering tree.

fineRADstructure itself (RADpainter + fineSTRUCTURE) is run outside R. This
script assumes RADpainter has already been run on the GoM10 and Sub VCF
subsets (produced in Section 1 below) to generate `.chunks.out`,
`.mcmc.xml`, and `.mcmcTree.xml` files, and that `FinestructureLibrary.R`
and `fineRADstructurePlot.R` (from the fineRADstructure distribution) are
available locally.

**A note on `fineRADstructurePlot.R`:** this external script must be manually
edited to point at the correct dataset's `chunkfile`/`mcmcfile`/`treefile`
before each `source()` call in this document — it does not take arguments.
Comments at each `source()` call indicate which dataset it must be pointing
to at that step.

## Requirements

R packages used across this script:
vcfR, tidyverse (dplyr, tidyr, readr, forcats), reshape2, viridis, pheatmap, scales

External tools / scripts:
fineRADstructure (RADpainter, fineSTRUCTURE) — Malinsky et al. 2018
FinestructureLibrary.R, fineRADstructurePlot.R


## Required input files

- `swiftia_snps_90_cov_maf_bi.recode.vcf` — filtered biallelic SNP VCF (also
  archived on Figshare)
- `pop_code_GoM10.Rdata`, `pop_code_sub.Rdata` — population code tables
- `popmap.txt` — two-column sample-to-population map (no header)
- `se_metadata_neutral_sub` — metadata including `SITE` and `sample_id`,
  covering all individuals in both datasets

## Outputs

**Figures (PDF/PNG):** west-east ordered population co-ancestry matrices
(GoM10, Sub); Sub individual co-ancestry heatmap (west-east ordered);
dendrogram-clustered individual co-ancestry heatmaps (GoM10, Sub)

**Tables (CSV):** individual and population-averaged co-ancestry matrices for
GoM10 and Sub.

## Notes

- **Directional convention:** in all co-ancestry matrices, rows = Recipient,
  columns = Donor.
- **Edisto sample ID mangling:** R's automatic column-name handling converts
  hyphenated Edisto sample IDs (e.g. `14-v-18-1-001`) to `X14.v.18.1.001`
  when they're read in as row/column names. The fix
  (`gsub("^X", "", gsub("\\.", "-", ...))`) must be applied immediately after
  each `source("fineRADstructurePlot.R")` call for the Sub dataset — it is
  not needed for GoM10, whose sample IDs are already in `SH*` format.

# Setup

```{r setup, message=FALSE}
library(vcfR)
library(tidyverse)
library(reshape2)
library(viridis)
library(pheatmap)
library(scales)
```

```{r base-dir}

base_dir <- "file/path"
setwd(base_dir)
dir.create("fineRADfigures", showWarnings = FALSE)
```

# 1. Export GoM10 and Sub VCF subsets for RADpainter

```{r vcf-subset-export}
vcf_full <- read.vcfR("swiftia_snps_90_cov_maf_bi.recode.vcf")

load("pop_code_GoM10.Rdata")
load("pop_code_sub.Rdata")

gom10_samples <- rownames(pop_code_GoM10)
sub_samples   <- rownames(pop_code_sub)

# write.vcf always produces .gz
vcf_gom10 <- vcf_full[, c(1, which(colnames(vcf_full@gt)[-1] %in% gom10_samples) + 1)]
vcfR::write.vcf(vcf_gom10, "swiftia_snps_GoM10.vcf.gz")

vcf_sub <- vcf_full[, c(1, which(colnames(vcf_full@gt)[-1] %in% sub_samples) + 1)]
vcfR::write.vcf(vcf_sub, "swiftia_snps_sub.vcf.gz")

cat("GoM10:", ncol(vcf_gom10@gt) - 1, "samples\n")
cat("Sub:  ", ncol(vcf_sub@gt) - 1,   "samples\n")
```

Run RADpainter on each `.vcf.gz` outside R to generate the `.chunks.out`,
`.mcmc.xml`, and `.mcmcTree.xml` files needed below.

# 2. Population-averaged co-ancestry matrices (west-to-east order)

## 2.1 GoM10 setup and checks
BEFORE RUNNING: edit `fineRADstructurePlot.R` to point to the GoM10 output
files (`chunkfile <- "swiftia_GoM10_chunks.out"`, etc.; `analysisName <-
"swiftia_GoM10"`).

```{r shared-pop-order}
WE_order_GoM10 <- c("EFGB", "Elvers", "McGrail", "Bouma",
                     "Alderdice", "Diaphus", "FTR", "AAR", "RTR")
WE_order_sub   <- c("EFGB", "Elvers", "McGrail", "Bouma",
                     "Alderdice", "Diaphus", "FTR", "AAR", "RTR", "Edisto")
```


```{r gom10-finerad-setup}
popmap <- read.table("popmap.txt", header = FALSE,
                      col.names = c("SAMPLE", "POP"),
                      stringsAsFactors = FALSE)
rownames(popmap) <- popmap$SAMPLE
cat("popmap loaded:", nrow(popmap), "samples,",
    length(unique(popmap$POP)), "populations\n")

# Produces: dataraw, datamatrix, popmeanmatrix, mapstatelist, tdend, popdend
source("FinestructureLibrary.R")
source("fineRADstructurePlot.R")

# CHECK 1: dataraw dimensions
cat("\n--- CHECK 1: dataraw dimensions ---\n")
cat("Expected: 158 x 158\n")
cat("Actual:  ", nrow(dataraw), "x", ncol(dataraw), "\n")
stopifnot(nrow(dataraw) == 158, ncol(dataraw) == 158)

# CHECK 2: population assignment (GoM10 IDs are SH* format; name mangling
# is not expected, but verify)
cat("\n--- CHECK 2: Population assignment ---\n")
ind_names <- rownames(dataraw)
ind_pops  <- popmap[ind_names, "POP"]
cat("NA population assignments:", sum(is.na(ind_pops)), "(expected 0)\n")
if (sum(is.na(ind_pops)) > 0) {
  rownames(dataraw) <- gsub("^X", "", gsub("\\.", "-", rownames(dataraw)))
  colnames(dataraw) <- gsub("^X", "", gsub("\\.", "-", colnames(dataraw)))
  ind_names <- rownames(dataraw)
  ind_pops  <- popmap[ind_names, "POP"]
  cat("After fix — NA assignments:", sum(is.na(ind_pops)), "\n")
  stopifnot(sum(is.na(ind_pops)) == 0)
}

# CHECK 3: all expected populations present
cat("\n--- CHECK 3: Populations present ---\n")
pops_found <- sort(unique(ind_pops))
pops_missing <- setdiff(WE_order_GoM10, pops_found)
cat("Populations found:", paste(pops_found, collapse = ", "), "\n")
if (length(pops_missing) > 0) {
  warning("Missing populations: ", paste(pops_missing, collapse = ", "))
} else {
  cat("All 9 expected populations present\n")
}
```

## 2.2 GoM10 population co-ancestry matrix (computation)
```{r gom10-pop-matrix-compute}
pops_unique <- WE_order_GoM10[WE_order_GoM10 %in% unique(ind_pops)]
pop_mat <- matrix(NA, nrow = length(pops_unique), ncol = length(pops_unique),
                   dimnames = list(pops_unique, pops_unique))

for (i in pops_unique) {
  for (j in pops_unique) {
    rows_i <- ind_names[ind_pops == i]
    cols_j <- ind_names[ind_pops == j]
    if (i == j) {
      sub_m <- dataraw[rows_i, cols_j, drop = FALSE]
      diag(sub_m) <- NA   # exclude self-comparisons
      pop_mat[i, j] <- mean(sub_m, na.rm = TRUE)
    } else {
      pop_mat[i, j] <- mean(dataraw[rows_i, cols_j, drop = FALSE], na.rm = TRUE)
    }
  }
}

# CHECK 4: matrix completeness
cat("\n--- CHECK 4: Population matrix completeness ---\n")
cat("Off-diagonal NAs:", sum(is.na(pop_mat[row(pop_mat) != col(pop_mat)])),
    "(expected 0)\n")

write.csv(round(pop_mat, 3), "swiftia_GoM10_pop_coancestry.csv")
cat("\nGoM10 population co-ancestry matrix (west to east):\n")
print(round(pop_mat, 2))

pop_mat_GoM10 <- pop_mat
```

## 2.3 Plotting function and final population-matrix figures
```{r plot-pop-coancestry-function}
plot_pop_coancestry <- function(pop_mat, pop_order, dataset_label,
                                 n_ind, n_snps, outfile_prefix) {

  pops <- pop_order[pop_order %in% rownames(pop_mat)]

  mat_plot <- pop_mat[pops, pops]
  diag(mat_plot) <- NA

  melted <- reshape2::melt(mat_plot)
  names(melted) <- c("Recipient", "Donor", "Coancestry")
  melted$Recipient <- factor(melted$Recipient, levels = pops)
  melted$Donor     <- factor(melted$Donor,     levels = pops)

  clim <- range(melted$Coancestry, na.rm = TRUE)

  fig <- ggplot(melted, aes(x = Donor, y = Recipient, fill = Coancestry)) +
    geom_tile(color = "grey90", linewidth = 0.4) +
    geom_tile(data = melted[is.na(melted$Coancestry), ],
              fill = "grey85", color = "grey90") +
    geom_text(data = melted[!is.na(melted$Coancestry), ],
              aes(label = round(Coancestry, 1)),
              color = "white", size = 3.2, fontface = "bold") +
    scale_fill_viridis_c(
      option   = "magma",
      limits   = clim,
      na.value = "grey85",
      name     = "Mean\nco-ancestry\n(chunk counts)"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      axis.text.x     = element_text(angle = 45, hjust = 1, face = "bold", size = 10),
      axis.text.y     = element_text(face = "bold", size = 10),
      axis.title      = element_text(face = "bold", size = 11),
      panel.grid      = element_blank(),
      legend.position = "right",
      legend.title    = element_text(size = 10),
      plot.title      = element_text(hjust = 0.5, face = "bold", size = 13),
      plot.subtitle   = element_text(hjust = 0.5, size = 10, colour = "grey40"),
      plot.caption    = element_text(size = 8, colour = "grey50", hjust = 0)
    ) +
    labs(
      title    = bquote(italic("Swiftia exserta") ~
                        "\u2014 Population co-ancestry" ~ .(paste0("(", dataset_label, ")"))),
      subtitle = "Mean RADpainter chunk counts between population pairs (west to east)\nDiagonal (within-population) excluded",
      x        = "Donor population",
      y        = "Recipient population",
      caption  = paste0("n=", n_ind, " individuals, ", length(pops),
                        " populations, ", n_snps, " neutral SNPs\n",
                        "fineRADstructure (Malinsky et al. 2018); ",
                        "loci reordered by LD prior to painting")
    )

  ggsave(paste0("figures/", outfile_prefix, ".pdf"), fig, width = 8, height = 7, dpi = 300)
  ggsave(paste0("figures/", outfile_prefix, ".png"), fig, width = 8, height = 7, dpi = 300)
  cat("Saved: figures/", outfile_prefix, ".pdf/png\n", sep = "")
  print(fig)
  return(fig)
}
```

```{r gom10-pop-matrix-plot}
fig_pop_GoM10 <- plot_pop_coancestry(
  pop_mat        = pop_mat_GoM10,
  pop_order      = WE_order_GoM10,
  dataset_label  = "GoM10",
  n_ind          = 158,
  n_snps         = "16,609",
  outfile_prefix = "Fig_coancestry_GoM10_population_matrix"
)
```

## 2.4 Sub individual heatmap and population matrix

BEFORE RUNNING: edit `fineRADstructurePlot.R` to point to the Sub output
files, then re-source (see CHECK below — expects `dataraw` to become 174x174).
```{r sub-pop-and-individual-matrices}
source("fineRADstructurePlot.R")

cat("\n=== dataraw check ===\n")
cat("Current dimensions:", dim(dataraw), "\n")
cat("For Sub heatmap need: 174 x 174\n")
stopifnot(nrow(dataraw) == 174)

# Edisto sample ID fix — required immediately after sourcing for Sub (see Notes above)
rownames(dataraw) <- gsub("^X", "", gsub("\\.", "-", rownames(dataraw)))
colnames(dataraw) <- gsub("^X", "", gsub("\\.", "-", colnames(dataraw)))

ind_names_sub <- rownames(dataraw)
ind_pops_sub  <- popmap[ind_names_sub, "POP"]
stopifnot(sum(is.na(ind_pops_sub)) == 0)

pop_factor_sub   <- factor(ind_pops_sub, levels = WE_order_sub)
ind_order_sub    <- order(pop_factor_sub)
mat_ordered_sub  <- dataraw[ind_order_sub, ind_order_sub]
pops_ordered_sub <- ind_pops_sub[ind_order_sub]

pop_breaks_sub <- cumsum(table(factor(pops_ordered_sub, levels = WE_order_sub))) + 0.5
pop_mids_sub   <- c(0, pop_breaks_sub[-length(pop_breaks_sub)]) +
                  diff(c(0, pop_breaks_sub)) / 2

# Off-diagonal color scale (1st-99th percentile, avoids diagonal dominance)
mat_offdiag_sub       <- mat_ordered_sub
diag(mat_offdiag_sub) <- NA
min_val_sub <- quantile(as.vector(mat_offdiag_sub), 0.01, na.rm = TRUE)
cap_val_sub <- quantile(as.vector(mat_offdiag_sub), 0.99, na.rm = TRUE)
cat("Off-diagonal color scale: min =", round(min_val_sub, 3),
    "| cap (99th pct) =", round(cap_val_sub, 3), "\n")

x_break_indices_sub <- pmax(1, pmin(round(pop_mids_sub), nrow(mat_ordered_sub)))
x_breaks_sub        <- rownames(mat_ordered_sub)[x_break_indices_sub]

mat_df_sub <- as.data.frame(mat_ordered_sub)
mat_df_sub$Recipient <- factor(rownames(mat_ordered_sub), levels = rownames(mat_ordered_sub))
melted_sub <- reshape2::melt(mat_df_sub, id.vars = "Recipient",
                              variable.name = "Donor", value.name = "Coancestry")

# --- Sub individual heatmap (west-to-east) ---
fig2 <- ggplot(melted_sub, aes(x = Donor, y = Recipient, fill = Coancestry)) +
  geom_tile() +
  geom_tile(data = melted_sub[as.integer(melted_sub$Donor) == as.integer(melted_sub$Recipient), ],
            fill = "grey80") +
  geom_vline(xintercept = pop_breaks_sub, colour = "white", linewidth = 0.5) +
  geom_hline(yintercept = nrow(mat_ordered_sub) - pop_breaks_sub + 1,
             colour = "white", linewidth = 0.5) +
  scale_fill_viridis_c(
    option   = "magma",
    limits   = c(min_val_sub, cap_val_sub),
    oob      = scales::squish,
    name     = "Co-ancestry\n(chunk counts)"
  ) +
  scale_x_discrete(breaks = x_breaks_sub, labels = WE_order_sub) +
  scale_y_discrete(breaks = rownames(mat_ordered_sub)[x_break_indices_sub], labels = WE_order_sub) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1, face = "bold", size = 10),
    axis.text.y     = element_text(face = "bold", size = 10),
    axis.title      = element_text(face = "bold", size = 11),
    panel.grid      = element_blank(),
    legend.position = "right",
    legend.title    = element_text(size = 10),
    plot.title      = element_text(hjust = 0.5, face = "bold", size = 13),
    plot.subtitle   = element_text(hjust = 0.5, size = 10, colour = "grey40"),
    plot.caption    = element_text(size = 8, colour = "grey50", hjust = 0),
    plot.margin     = margin(t = 10, r = 20, b = 40, l = 10, unit = "pt")
  ) +
  labs(
    title    = expression(italic("Swiftia exserta") ~ "\u2014 Co-ancestry matrix (Sub)"),
    subtitle = paste0("n=174 individuals ordered west to east by population\n",
                      "Color scale restricted to off-diagonal range; ",
                      "white lines denote population boundaries"),
    x        = "Donor individual",
    y        = "Recipient individual",
    caption  = paste0("10 populations (9 Gulf of Mexico + Edisto), 16,688 neutral SNPs\n",
                      "fineRADstructure (Malinsky et al. 2018); loci reordered by LD prior to painting")
  )

ggsave("figures/Fig_coancestry_Sub_individual_heatmap.pdf", fig2, width = 10, height = 9, dpi = 300)
ggsave("figures/Fig_coancestry_Sub_individual_heatmap.png", fig2, width = 10, height = 9, dpi = 300)
cat("Saved: figures/Fig_coancestry_Sub_individual_heatmap.pdf/png\n")
print(fig2)

# --- Sub population-averaged matrix ---
pops_unique_sub <- WE_order_sub[WE_order_sub %in% unique(ind_pops_sub)]
pop_mat_sub <- matrix(NA, nrow = length(pops_unique_sub), ncol = length(pops_unique_sub),
                       dimnames = list(pops_unique_sub, pops_unique_sub))
for (i in pops_unique_sub) {
  for (j in pops_unique_sub) {
    rows_i <- ind_names_sub[ind_pops_sub == i]
    cols_j <- ind_names_sub[ind_pops_sub == j]
    if (i == j) {
      sub_m <- dataraw[rows_i, cols_j, drop = FALSE]
      diag(sub_m) <- NA
      pop_mat_sub[i, j] <- mean(sub_m, na.rm = TRUE)
    } else {
      pop_mat_sub[i, j] <- mean(dataraw[rows_i, cols_j, drop = FALSE], na.rm = TRUE)
    }
  }
}
write.csv(round(pop_mat_sub, 3), "swiftia_sub_pop_coancestry.csv")
cat("\nSub population co-ancestry matrix (west to east):\n")
print(round(pop_mat_sub, 2))

fig_pop_Sub <- plot_pop_coancestry(
  pop_mat        = pop_mat_sub,
  pop_order      = WE_order_sub,
  dataset_label  = "Sub",
  n_ind          = 174,
  n_snps         = "16,688",
  outfile_prefix = "Fig_coancestry_Sub_population_matrix"
)
```

# 3. Individual-level co-ancestry heatmaps (dendrogram-clustered)

Individuals here are ordered by the fineRADstructure hierarchical clustering
tree (`tdend`) rather than west-to-east geography — this is the complementary
view to Section 2's geography-ordered matrices, showing whether the
co-ancestry-based clustering recovers the same structure independent of
imposed population order.
```{r dendrogram-setup}
pop_colors_GoM10 <- c(
  "EFGB"      = "#FF2E00",
  "McGrail"   = "#F78C61",
  "Elvers"    = "#FFB812",
  "Bouma"     = "#6F8E00",
  "Alderdice" = "#1B6758",
  "Diaphus"   = "#B0FFAD",
  "FTR"       = "#45C2E3",
  "AAR"       = "#000872",
  "RTR"       = "#F5C7FF"
)
pop_colors_sub <- c(pop_colors_GoM10, "Edisto" = "#6E0A7C")

# Population lookup: sample_id -> site (covers individuals in both datasets)
pop_lookup <- setNames(se_metadata_neutral_sub$SITE, se_metadata_neutral_sub$sample_id)
```

## 3.1 Sub — dendrogram-clustered heatmap

BEFORE RUNNING: edit `fineRADstructurePlot.R` to point to the Sub output
files (`swiftia_sub_chunks.mcmc.xml`, etc.).

```{r sub-dendrogram-heatmap}
source("fineRADstructurePlot.R")

# Edisto sample ID fix — applied immediately after sourcing (see Notes above)
rownames(dataraw) <- gsub("^X", "", gsub("\\.", "-", rownames(dataraw)))
colnames(dataraw) <- gsub("^X", "", gsub("\\.", "-", colnames(dataraw)))

write.csv(dataraw, file = "swiftia_sub_ind_coancestry.csv", row.names = TRUE)

# Symmetrize and remove diagonal
datamat <- (dataraw + t(dataraw)) / 2
diag(datamat) <- NA

# Fix Edisto labels in the clustering tree, convert to hclust
tdend <- dendrapply(tdend, function(n) {
  if (!is.null(attr(n, "label"))) {
    attr(n, "label") <- gsub("^X", "", gsub("\\.", "-", attr(n, "label")))
  }
  n
})
tdend_hclust <- as.hclust(tdend)

stopifnot(all(tdend_hclust$labels %in% rownames(datamat)))
stopifnot(all(rownames(datamat) %in% tdend_hclust$labels))

datamat_ordered <- datamat[tdend_hclust$labels, tdend_hclust$labels]

pops_sub <- pop_lookup[tdend_hclust$labels]
ann_sub <- data.frame(
  Population = factor(pops_sub, levels = WE_order_sub),
  row.names  = tdend_hclust$labels
)

pdf("figures/fineRAD_ind_heatmap_sub_clustered_annotated_final.pdf", width = 11, height = 10)
pheatmap(
  datamat_ordered,
  color                = magma(100),
  breaks               = seq(62, 74, length.out = 101),
  cluster_rows         = tdend_hclust,
  cluster_cols         = tdend_hclust,
  show_rownames        = FALSE,
  show_colnames        = FALSE,
  annotation_row       = ann_sub,
  annotation_col       = ann_sub,
  annotation_colors    = list(Population = pop_colors_sub),
  annotation_names_row = FALSE,
  annotation_names_col = FALSE,
  border_color         = NA,
  na_col               = "white",
  treeheight_row       = 50,
  treeheight_col       = 50,
  main                 = "Individual co-ancestry \u2014 Sub (clustered)"
)
dev.off()
```

## 3.2 GoM10 — dendrogram-clustered heatmap

BEFORE RUNNING: edit `fineRADstructurePlot.R` to point to the GoM10 output
files, then re-source.

```{r gom10-dendrogram-heatmap}
source("fineRADstructurePlot.R")
# No Edisto name fix needed for GoM10 (SH* IDs, no mangling)

write.csv(dataraw, file = "swiftia_GoM10_ind_coancestry.csv", row.names = TRUE)

datamat_GoM10 <- (dataraw + t(dataraw)) / 2
diag(datamat_GoM10) <- NA

tdend_hclust_GoM10 <- as.hclust(tdend)

stopifnot(all(tdend_hclust_GoM10$labels %in% rownames(datamat_GoM10)))
stopifnot(all(rownames(datamat_GoM10) %in% tdend_hclust_GoM10$labels))

datamat_GoM10_ordered <- datamat_GoM10[tdend_hclust_GoM10$labels, tdend_hclust_GoM10$labels]

pops_GoM10 <- pop_lookup[tdend_hclust_GoM10$labels]
ann_GoM10 <- data.frame(
  Population = factor(pops_GoM10, levels = WE_order_GoM10),
  row.names  = tdend_hclust_GoM10$labels
)

# Off-diagonal range — adjust breaks below if this summary changes materially
offdiag_GoM10 <- datamat_GoM10[upper.tri(datamat_GoM10)]
summary(offdiag_GoM10)

pdf("figures/fineRAD_ind_heatmap_GoM10_clustered_annotated_final.pdf", width = 10, height = 9)
pheatmap(
  datamat_GoM10_ordered,
  color                = magma(100),
  breaks               = seq(68, 78, length.out = 101),
  cluster_rows         = tdend_hclust_GoM10,
  cluster_cols         = tdend_hclust_GoM10,
  show_rownames        = FALSE,
  show_colnames        = FALSE,
  annotation_row       = ann_GoM10,
  annotation_col       = ann_GoM10,
  annotation_colors    = list(Population = pop_colors_GoM10),
  annotation_names_row = FALSE,
  annotation_names_col = FALSE,
  border_color         = NA,
  na_col               = "white",
  treeheight_row       = 50,
  treeheight_col       = 50,
  main                 = "Individual co-ancestry \u2014 GoM10 (clustered)"
)
dev.off()
```


