---
title: "05_Plot_Physical_Connectivity"
author: "Nicole Pittoors"
date: "2026-08-25"
output: html_document
---

# Overview

This script plots the biophysical larval dispersal model output (Ichthyop/CROCO,
provided by Luisa) for *Swiftia exserta* in the northern Gulf of Mexico. It
produces two figures:

1. **Connectivity matrix** — full 12-population ensemble matrix (release x
   recruitment zone), showing larval settlement proportions (%) with year(s)
   of connection (2014 / 2015 / Both) indicated by point shape.
2. **Connectivity map** — GoM10 (9-population) spatial network map, showing
   directed larval exchange arrows over a Gulf of Mexico basemap. Arrows
   >=0.1% transport are magma-colored by exchange strength; trace exchange
   (<0.1%, greater than zero) is shown as thin dashed arrows.

This script is standalone — it depends only on the raw connectivity CSVs, not
on any object from the population genetics scripts (`01`-`04`).

## Requirements
tidyverse (dplyr, tidyr, readr, purrr, forcats), viridis, scales, marmap, maps

## Required input files

- `Se_popgen_Connectivity_matrix.csv` — larval settlement proportions (%),
  release zones (rows) x recruitment zones (columns)
- `Se_popgen_Connectivity_matrix_years.csv` — same dimensions; release year(s)
  ("2014", "2015", "Both") per cell, NA where no modeled connection exists

## Outputs

**Figures (PDF/PNG):** `phys_connectivity_matrix` (full 12-population
ensemble matrix); `swiftia_GoM10_connectivity_map` (9-population directed
network map).

## Notes

- **RTR fix:** `Se_popgen_Connectivity_matrix.csv` has RTR's row label read in
  as `NA` rather than `"RTR"` — this is fixed programmatically in Section 2
  rather than by editing the source CSV, so the fix is visible and
  reproducible. If RTR is also absent as a *column* (no other site recruits
  from RTR in the model), a zero-column is added rather than leaving it
  missing, so RTR still appears in the GoM10 subset matrix.
- **`<0.1` values:** the source CSV uses `"<0.1"` as a floor label for very
  small but nonzero connectivity. This is treated numerically as `0.05` for
  fill-color purposes in the matrix figure; for the map figure, `<0.1` edges
  are kept in a separate "trace exchange" layer rather than dropped, so the
  reader can see where a modeled connection exists at all versus where it's
  strong enough to plot with an arrow.
- **GoM10 subset:** the map uses only the 9 GoM10 populations (drops WFGB,
  Bright, Geyer — the small-n sites excluded from GoM10 throughout the rest
  of the pipeline), matching the population genetics side for direct
  comparison in the manuscript.


# Setup

```{r setup, message=FALSE}
library(tidyverse)
library(viridis)
library(scales)
library(marmap)
library(maps)
```

```{r base-dir}
base_dir <- "your/base/path"
setwd(base_dir)
dir.create("figures", showWarnings = FALSE)
```

# 1. Import connectivity data

```{r import-connectivity-data}
val_raw  <- read_csv("Se_popgen_Connectivity_matrix.csv",       show_col_types = FALSE)
year_raw <- read_csv("Se_popgen_Connectivity_matrix_years.csv", show_col_types = FALSE)

parse_matrix <- function(raw) {
  df         <- as.data.frame(raw)
  col_labels <- as.character(unlist(df[nrow(df), ]))
  row_labels <- as.character(df[[1]])
  row_labels <- row_labels[!is.na(row_labels) & row_labels != "NA"]
  data_block <- df[-nrow(df), -1]
  colnames(data_block) <- col_labels[-1]
  rownames(data_block) <- row_labels
  data_block
}

val_data  <- parse_matrix(val_raw)
year_data <- parse_matrix(year_raw)

recruit_order <- c("WFGB", "EFGB", "Bright", "Geyer", "Elvers",
                    "McGrail", "Bouma", "Alderdice", "Diaphus", "FTR", "AAR", "RTR")
release_order <- c("RTR", "AAR", "FTR", "Diaphus", "Alderdice", "Bouma",
                    "McGrail", "Elvers", "Geyer", "Bright", "EFGB", "WFGB")

val_data  <- val_data[release_order, recruit_order]
year_data <- year_data[release_order, recruit_order]
```

# 2. GoM10 subset and fixes (needed for the map figure only)

```{r gom10-subset-fixes}
GoM10_sites <- c("EFGB", "Elvers", "McGrail", "Bouma",
                  "Alderdice", "Diaphus", "FTR", "AAR", "RTR")

# Convert "<0.1" -> 0.05, everything else to numeric
larval_numeric <- val_data
larval_numeric[] <- lapply(val_data, function(x)
  ifelse(x == "<0.1", 0.05, suppressWarnings(as.numeric(x))))
larval_numeric <- as.data.frame(lapply(larval_numeric, as.numeric))
rownames(larval_numeric) <- rownames(val_data)

# Fix 1: RTR row label was read in as "NA" rather than "RTR"
rownames(larval_numeric)[rownames(larval_numeric) == "NA"] <- "RTR"

# Fix 2: add an RTR column of zeros if it's absent as a recruitment zone
# (i.e. RTR received no modeled larval input from any site)
if (!"RTR" %in% colnames(larval_numeric)) {
  larval_numeric[["RTR"]] <- 0
  cat("RTR column added as zeros (RTR received no modeled larval input)\n")
}

stopifnot(all(GoM10_sites %in% rownames(larval_numeric)))
stopifnot(all(GoM10_sites %in% colnames(larval_numeric)))

# Subset to GoM10 sites only (drops WFGB, Bright, Geyer)
larval_GoM10 <- larval_numeric[GoM10_sites, GoM10_sites]
larval_GoM10[is.na(larval_GoM10)] <- 0

cat("larval_GoM10 dims:", dim(larval_GoM10), "\n")
```

# 3. Figure 1 — Connectivity matrix (full ensemble, 12 populations)

```{r connectivity-matrix}
to_long <- function(mat, value_col) {
  mat %>%
    as.data.frame() %>%
    mutate(Release = rownames(.)) %>%
    pivot_longer(-Release, names_to = "Recruit", values_to = value_col)
}

df <- left_join(
  to_long(val_data,  "Value"),
  to_long(year_data, "Year"),
  by = c("Release", "Recruit")
) %>%
  filter(!is.na(Year), Year != "NA") %>%
  mutate(
    # "<0.1" is a display-only floor label; treated numerically as 0.05 for fill
    NumValue = ifelse(Value == "<0.1", 0.05, suppressWarnings(as.numeric(Value))),
    Label    = as.character(Value),
    Release  = factor(Release, levels = release_order),
    Recruit  = factor(Recruit, levels = recruit_order),
    Year     = factor(Year,    levels = c("2014", "2015", "Both"))
  )

p_connectivity <- ggplot(df, aes(x = Recruit, y = Release)) +
  geom_tile(aes(fill = NumValue), color = "white", linewidth = 0.5) +
  geom_hline(yintercept = seq(0.5, length(release_order) - 0.5, by = 1),
             color = "grey85", linewidth = 0.4) +
  geom_vline(xintercept = seq(0.5, length(recruit_order) - 0.5, by = 1),
             color = "grey85", linewidth = 0.4) +
  geom_text(aes(label = Label), color = "white", size = 2.8, fontface = "bold") +
  geom_point(
    aes(x = as.numeric(Recruit) + 0.32,
        y = as.numeric(fct_rev(Release)) + 0.32,
        shape = Year),
    color = "white", size = 1.8, stroke = 0.3
  ) +
  scale_fill_gradientn(
    colors = magma(100, begin = 0.10, end = 0.95),
    name   = "Connectivity\n(%)",
    guide  = guide_colorbar(barwidth = 0.8, barheight = 8,
                             title.position = "top", title.hjust = 0.5)
  ) +
  scale_shape_manual(values = c("2014" = 16, "2015" = 17, "Both" = 15), name = "Year") +
  scale_x_discrete(position = "bottom") +
  scale_y_discrete(limits = rev(release_order)) +
  labs(x = "Recruitment Zone", y = "Release Zone") +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x      = element_text(angle = 45, hjust = 1, vjust = 1, size = 9),
    axis.text.y      = element_text(size = 9),
    axis.title       = element_text(face = "bold", size = 11),
    panel.grid       = element_blank(),
    legend.position  = "right",
    legend.box       = "vertical",
    legend.spacing.y = unit(0.5, "cm"),
    legend.title     = element_text(face = "bold", size = 9),
    legend.text      = element_text(size = 9),
    plot.margin      = margin(8, 8, 8, 8)
  ) +
  coord_cartesian(clip = "off")

ggsave("figures/phys_connectivity_matrix.png", plot = p_connectivity,
       width = 8, height = 7, dpi = 300, bg = "white")
ggsave("figures/phys_connectivity_matrix.pdf", plot = p_connectivity,
       width = 8, height = 7, dpi = 300)

print(p_connectivity)
cat("Saved: figures/phys_connectivity_matrix.pdf/png\n")
```

# 4. Figure 2 prep — directed edges and basemap (GoM10, 9 populations)

```{r gom10-site-coords}
# GoM10 site coordinates (no WFGB, Bright, Geyer)
sites_df <- data.frame(
  site = GoM10_sites,
  lon  = c(-93.61300, -92.92279, -92.82152, -92.46734,
           -91.98353, -90.70327, -88.43804, -88.33845, -87.57561),
  lat  = c( 27.96824,  27.85359,  27.96214,  28.07165,
            28.07640,  28.08492,  29.23241,  29.25225,  29.43917)
)

# label offsets below are a uniform default. Tune label_hjust/label_vjust per site if any label overlaps the coastline or another point.
label_df <- sites_df %>%
  mutate(label_hjust = -0.15, label_vjust = -0.55)
```

```{r build-directed-edges}
# All nonzero directed larval edges (release -> recruit)
directed_larval_all <- data.frame()
for (i in seq_along(GoM10_sites)) {
  for (j in seq_along(GoM10_sites)) {
    if (i == j) next
    val <- as.numeric(larval_GoM10[GoM10_sites[i], GoM10_sites[j]])
    if (is.na(val) || val == 0) next
    directed_larval_all <- rbind(directed_larval_all, data.frame(
      from     = GoM10_sites[i],
      to       = GoM10_sites[j],
      larv_val = val
    ))
  }
}

cat("All nonzero directed larval edges:", nrow(directed_larval_all), "\n")
cat("Value range:", round(range(directed_larval_all$larv_val), 3), "\n")

# Main edges: >= 0.1% transport, magma-colored by strength (log scale)
directed_larval_B <- directed_larval_all %>%
  filter(larv_val >= 0.1) %>%
  left_join(sites_df, by = c("from" = "site")) %>%
  rename(from_lon = lon, from_lat = lat) %>%
  left_join(sites_df, by = c("to" = "site")) %>%
  rename(to_lon = lon, to_lat = lat) %>%
  mutate(
    curve = case_when(
      from == "Elvers"    & to == "McGrail"   ~ -0.25,
      from == "McGrail"   & to == "EFGB"      ~  0.30,
      from == "McGrail"   & to == "Bouma"     ~  0.35,
      from == "McGrail"   & to == "Alderdice" ~  0.20,
      from == "Bouma"     & to == "EFGB"      ~ -0.20,
      from == "Bouma"     & to == "McGrail"   ~ -0.35,
      from == "Alderdice" & to == "EFGB"      ~ -0.30,
      from == "Alderdice" & to == "McGrail"   ~  0.25,
      from == "Alderdice" & to == "Bouma"     ~ -0.20,
      from == "FTR"       & to == "AAR"       ~  0.80,
      from == "AAR"       & to == "FTR"       ~ -0.80,
      TRUE ~ 0.20
    ),
    palette_pos = scales::rescale(log1p(larv_val), to = c(0.15, 0.95),
                                   from = log1p(c(0.1, 7.3))),
    magma_col = mapply(function(p) viridis::magma(1, begin = p, end = p), palette_pos)
  )

# Trace edges: nonzero but < 0.1% transport — shown as thin dashed arrows
# curve is a uniform default here (0.15) rather than pair-tuned,adjust
# per-pair (as done for directed_larval_B above) if any trace arrows overlap.
directed_larval_trace <- directed_larval_all %>%
  filter(larv_val < 0.1) %>%
  left_join(sites_df, by = c("from" = "site")) %>%
  rename(from_lon = lon, from_lat = lat) %>%
  left_join(sites_df, by = c("to" = "site")) %>%
  rename(to_lon = lon, to_lat = lat) %>%
  mutate(curve = 0.15)

cat("Main edges (>=0.1%):", nrow(directed_larval_B), "\n")
cat("Trace edges (<0.1%, >0):", nrow(directed_larval_trace), "\n")
```

```{r gom10-basemap}
gulf_bath <- getNOAA.bathy(lon1 = -95, lon2 = -87,
                            lat1 = 27,  lat2 = 30.5,
                            resolution = 4)
bath_df     <- fortify(gulf_bath)
gulf_states <- map_data("state",
                         region = c("texas", "louisiana", "mississippi", "alabama", "florida"))
```

# 5. Figure 2 — Connectivity map (GoM10, directed larval exchange)

```{r fig-connectivity-map}
connectivity_map <- ggplot() +

  geom_rect(aes(xmin = -95.2, xmax = -86.5, ymin = 26.8, ymax = 30.8),
            fill = "#D1EEFC") +

  geom_contour(data = bath_df,
               aes(x = x, y = y, z = z),
               color = "grey60", linewidth = 0.2,
               breaks = c(-500, -1000, -2000, -3000)) +

  geom_polygon(data = gulf_states,
               aes(x = long, y = lat, group = group),
               fill = "grey88", color = "grey50", linewidth = 0.3) +

  # LAYER 1: trace arrows (<0.1%) — thin, dashed, grey, drawn first
  {
    purrr::pmap(
      list(directed_larval_trace$from_lon, directed_larval_trace$from_lat,
           directed_larval_trace$to_lon,   directed_larval_trace$to_lat,
           directed_larval_trace$curve),
      function(x, y, xend, yend, curv) {
        geom_curve(aes(x = x, y = y, xend = xend, yend = yend),
                   curvature = curv,
                   color     = "grey60",
                   linewidth = 0.3,
                   linetype  = "dashed",
                   alpha     = 0.5,
                   arrow     = arrow(length = unit(0.08, "cm"), type = "closed", angle = 15))
      }
    )
  } +

  # LAYER 2: main arrows (>=0.1%) — magma colored, drawn on top
  {
    purrr::pmap(
      list(directed_larval_B$from_lon, directed_larval_B$from_lat,
           directed_larval_B$to_lon,   directed_larval_B$to_lat,
           directed_larval_B$curve,    directed_larval_B$magma_col),
      function(x, y, xend, yend, curv, col) {
        geom_curve(aes(x = x, y = y, xend = xend, yend = yend),
                   curvature = curv,
                   color     = col,
                   linewidth = 1.0,
                   alpha     = 0.9,
                   arrow     = arrow(length = unit(0.12, "cm"), type = "closed", angle = 15))
      }
    )
  } +

  # Invisible points to drive the magma colorbar
  geom_point(data = directed_larval_B,
             aes(x = from_lon, y = from_lat, color = larv_val),
             shape = NA) +

  scale_color_viridis_c(
    option = "magma",
    trans  = "log1p",
    breaks = c(0.1, 0.5, 1.0, 2.0, 5.0, 7.3),
    labels = c("0.1", "0.5", "1.0", "2.0", "5.0", "7.3"),
    limits = c(0.1, 7.3),
    begin  = 0.15,
    end    = 0.95,
    name   = "Larval transport\n(%)",
    guide  = guide_colorbar(barwidth = 0.8, barheight = 5,
                             title.position = "top", title.hjust = 0.5)
  ) +

  # Dummy segment so the trace-arrow style gets its own legend entry
  geom_segment(
    data = data.frame(x = NA_real_, y = NA_real_, xend = NA_real_, yend = NA_real_,
                       lab = "Trace exchange\n(<0.1%)"),
    aes(x = x, y = y, xend = xend, yend = yend, linetype = lab)
  ) +
  scale_linetype_manual(
    values = c("Trace exchange\n(<0.1%)" = "dashed"),
    name   = NULL,
    guide  = guide_legend(override.aes = list(color = "grey60", linewidth = 0.5, alpha = 0.6))
  ) +

  geom_point(data = sites_df,
             aes(x = lon, y = lat),
             size = 3.5, shape = 21,
             fill = "black", color = "white", stroke = 0.7) +

  geom_text(data = label_df,
            aes(x = lon, y = lat, label = site,
                hjust = label_hjust, vjust = label_vjust),
            size = 3.0, color = "grey20") +

  scale_x_continuous(breaks = seq(-95, -87, by = 1)) +
  scale_y_continuous(breaks = seq(27, 30, by = 1)) +
  coord_fixed(xlim = c(-95, -86.5), ylim = c(27, 30.4), ratio = 1) +

  theme_minimal(base_size = 11) +
  theme(
    panel.grid        = element_blank(),
    panel.border      = element_rect(color = "black", fill = NA, linewidth = 0.8),
    axis.ticks        = element_line(color = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    axis.title        = element_text(size = 11),
    legend.position   = "right",
    legend.title      = element_text(size = 9, face = "bold"),
    legend.text       = element_text(size = 8),
    legend.spacing.y  = unit(0.5, "cm")
  ) +
  labs(x = "Longitude", y = "Latitude")

print(connectivity_map)

ggsave("figures/swiftia_GoM10_connectivity_map.pdf",
       connectivity_map, width = 10, height = 6, dpi = 300)
ggsave("figures/swiftia_GoM10_connectivity_map.png",
       connectivity_map, width = 10, height = 6, dpi = 300)
```
