---
title: "01_Data_Setup_p"
author: "Nicole Pittoors"
date: "2026-08-25"
output: html_document
---

# Set directories, load metadata
```{r set directories}
main_dir <- "Your/Working/Directory"
output_dir <- "Your/Output/Directory"
setwd(main_dir)

# Load sample metadata (SITE, LAT, LONG, DEPTH per individual; also archived
# on Figshare as se_metadata_all_snps.csv). row.names = 1 restores sample IDs
# as rownames, matching how the rest of this script indexes se_metadata.
se_metadata <- read.csv("se_metadata_all_snps.csv", row.names = 1, stringsAsFactors = FALSE)
```


# Load Libraries
```{r load libraries}
library(vcfR)
library(dplyr)
library(adegenet)
library(poppr)
library(dartR)
library(qgraph)
library(hierfstat)
library(ggplot2)
library(pegas)
library(radiator)
library(umap)
library(assigner)
library(tidyr)
library(ggdendro)
library(pheatmap)
library(tidyverse)
library(viridis)
```

# Define colors and sites
```{r define colors, sites}
# Matches site for ARMS paper as best as possible
c13_sites <- c("Alderdice" = "#60714E",      
               "Geyer" = "#015951",
               "Elvers" = "#3B8C6E",
               "Bright" = "#B0DBBB",
               "Bouma" = "#5BBFB0",
               "Diaphus" = "#9DC9D3",
               "FTR" = "#408BA8",
               "WFGB" = "#0D3D59",
               "EFGB" = "#A8B1D6",
               "AAR" = "#FEBC2C",
               "McGrail" = "#FC8D2B",
               "RTR" = "#F1531B",
               "Edisto"= "#802A5A")

c_13sites_WE <- c("WFGB" = "#F1531B",  
  "EFGB" = "#FC8D2B",  
  "Bright" = "#FEBC2C",  
  "Geyer" = "#802A5A",  
  "McGrail" = "#A8B1D6",  
  "Elvers" = "#9DC9D3",  
  "Bouma" = "#408BA8",  
  "Alderdice" = "#0D3D59",  
  "Diaphus" = "#5BBFB0",  
  "FTR" = "#B0DBBB",  
  "AAR" = "#3B8C6E",  
  "RTR" = "#015951",  
  "Edisto" = "#60714E")

site_colors <- c(
  "WFGB" = "#700500",
  "EFGB" = "#FF2EBF",
  "Bright" = "#FF2E00",
  "Geyer" = "#FF7514",
  "McGrail" = "#F78C61",
  "Elvers" = "#FFB812",
  "Bouma" = "#6F8E00",
  "Alderdice" = "#1B6758",
  "Diaphus" = "#B0FFAD",
  "FTR" = "#45C2E3",
  "AAR" = "#000872",
  "RTR" = "#F5C7FF",
  "Edisto" = "#6E0A7C"
)

# Create a vector of site names in west-to-east order
W2E <- c('WFGB','EFGB','Bright','Geyer','McGrail','Elvers','Bouma','Alderdice','Diaphus','FTR','AAR','RTR','Edisto')

# Create color mapping for sites
sitecol <- data.frame(
  W2E = names(c13_sites),
  mypal = unname(c13_sites),
  stringsAsFactors = FALSE
)

# Arrange colors in west-to-east order
myCol.df <- sitecol[match(W2E, sitecol$W2E), ]
myCol <- myCol.df$mypal
```

=======================================================================
---------------------------Initial Filtering---------------------------
========================================================================
```{r create snps data}
vcf <- vcfR::read.vcfR("swiftia_snps_90_cov_maf_bi.recode.vcf")
vcf

# Transform vcf to genind
genind_all <- vcfR::vcfR2genind(vcf) #all SNP
# Export for figshare
save(genind_all,file='all_snps_genind.Rdata')

# Remove loci with too many missing (10% allowable missing data in this case)
filt.loci <- missingno(genind_all, type = "loci", cutoff = 0.10, quiet = FALSE, freq = FALSE)
filt.loci

# Remove individuals with too many missing (30% allowable missing data in this case)
filt.loci_ind <- missingno(filt.loci, type =  "geno", cutoff = 0.3, quiet = FALSE, freq = FALSE)

# Load pop info
pop <- read.table("popmap.txt", header=F, sep="\t", stringsAsFactors = TRUE)
colnames(pop)<-c('INDIVIDUALS','STRATA')
rownames(pop)<-pop$INDIVIDUALS

# Check for differences bewteen snp file and popmap file
ind_snps<-as.vector(indNames(filt.loci_ind), mode = 'character')
ind_pop<-as.vector(pop[,1], mode = 'character')
print(ind_snps %in% ind_pop)
setdiff(ind_snps,ind_pop)

# Exclude Edisto
#pop<-filter(pop, STRATA != 'Edisto')

#Add population info
ind<-pop[indNames(filt.loci_ind),]
ind<-na.omit(ind)
snps <- filt.loci_ind[ind$INDIVIDUALS]
snps@pop <- ind$STRATA
summary(snps@pop)

save(snps,file='snps.Rdata')
```

========================================================================
----------------------------Clone Analysis-----------------------------
========================================================================

```{r load snps}
load('snps.Rdata')
snps.gl<-gi2gl(snps)
gl2gds(snps.gl, outfile = "snps.gds", outpath = output_dir)
```

# Check visually for clones
First pass look before more stringent classification of clones
```{r visual check for clones}
pdf('./genetic_distances.pdf')
#AAR
#pairwise number of allelic differences between two individuals (diss.dist {poppr})
distgenDISS_AAR <- diss.dist(snps[pop='AAR'], percent = FALSE, mat = FALSE)
hist(distgenDISS_AAR)
save(distgenDISS_AAR, file = 'distgenDISS_AAR.Rdata')
#Visualize genetic distances  as network to detect clusters of highly related individuals (potential clones)
qgraph(1/distgenDISS_AAR, layout='spring', vsize=5, minimum = 0.00020)
#Alderdice
distgenDISS_ALD <- diss.dist(snps[pop='Alderdice'], percent = FALSE, mat = FALSE)
hist(distgenDISS_ALD)
save(distgenDISS_ALD, file = 'distgenDISS_ALD.Rdata')
qgraph(1/distgenDISS_ALD, layout='spring', vsize=5, minimum = 0.00020)
#Bouma
distgenDISS_BOU <- diss.dist(snps[pop='Bouma'], percent = FALSE, mat = FALSE)
hist(distgenDISS_BOU)
save(distgenDISS_BOU, file = 'distgenDISS_BOU.Rdata')
qgraph(1/distgenDISS_BOU, layout='spring', vsize=5, minimum = 0.00020)
#Bright
distgenDISS_BRI <- diss.dist(snps[pop='Bright'], percent = FALSE, mat = FALSE)
hist(distgenDISS_BRI)
save(distgenDISS_BRI, file = 'distgenDISS_BRI.Rdata')
qgraph(1/distgenDISS_BRI, layout='spring', vsize=5, minimum = 0.00020)
#Diaphus
distgenDISS_DIA <- diss.dist(snps[pop='Diaphus'], percent = FALSE, mat = FALSE)
hist(distgenDISS_DIA)
save(distgenDISS_DIA, file = 'distgenDISS_DIA.Rdata')
qgraph(1/distgenDISS_DIA, layout='spring', vsize=5, minimum = 0.00020)
#EFGB
distgenDISS_EFGB <- diss.dist(snps[pop='EFGB'], percent = FALSE, mat = FALSE)
hist(distgenDISS_EFGB)
save(distgenDISS_EFGB, file = 'distgenDISS_EFGB.Rdata')
qgraph(1/distgenDISS_EFGB, layout='spring', vsize=5, minimum = 0.00020)
#Edisto
distgenDISS_EDI <- diss.dist(snps[pop='Edisto'], percent = FALSE, mat = FALSE)
hist(distgenDISS_EDI)
save(distgenDISS_EDI, file = 'distgenDISS_EDI.Rdata')
qgraph(1/distgenDISS_EDI, layout='spring', vsize=5, minimum = 0.00020)
#Elvers
distgenDISS_ELV <- diss.dist(snps[pop='Elvers'], percent = FALSE, mat = FALSE)
hist(distgenDISS_ELV)
save(distgenDISS_ELV, file = 'distgenDISS_ELV.Rdata')
qgraph(1/distgenDISS_ELV, layout='spring', vsize=5, minimum = 0.00020)
#FTR
distgenDISS_FTR <- diss.dist(snps[pop='FTR'], percent = FALSE, mat = FALSE)
hist(distgenDISS_FTR)
save(distgenDISS_FTR, file = 'distgenDISS_FTR.Rdata')
qgraph(1/distgenDISS_FTR, layout='spring', vsize=5, minimum = 0.00020)
#Geyer
distgenDISS_GEY <- diss.dist(snps[pop='Geyer'], percent = FALSE, mat = FALSE)
hist(distgenDISS_GEY)
save(distgenDISS_GEY, file = 'distgenDISS_GEY.Rdata')
qgraph(1/distgenDISS_GEY, layout='spring', vsize=5, minimum = 0.00020)
#McGrail
distgenDISS_MCG <- diss.dist(snps[pop='McGrail'], percent = FALSE, mat = FALSE)
hist(distgenDISS_MCG)
save(distgenDISS_MCG, file = 'distgenDISS_MCG.Rdata')
qgraph(1/distgenDISS_MCG, layout='spring', vsize=5, minimum = 0.00020)
#RTR
distgenDISS_RTR <- diss.dist(snps[pop='RTR'], percent = FALSE, mat = FALSE)
hist(distgenDISS_RTR)
save(distgenDISS_RTR, file = 'distgenDISS_RTR.Rdata')
qgraph(1/distgenDISS_RTR, layout='spring', vsize=5, minimum = 0.00020)
#WFGB
distgenDISS_WFGB <- diss.dist(snps[pop='WFGB'], percent = FALSE, mat = FALSE)
hist(distgenDISS_WFGB)
save(distgenDISS_WFGB, file = 'distgenDISS_WFGB.Rdata')
qgraph(1/distgenDISS_WFGB, layout='spring', vsize=5, minimum = 0.00020)
dev.off()
```

# PLot all Histograms in one figure
```{r}
# Method 1: Using base R with par()
# This sets up a 4x3 grid of plots (adjust rows and columns as needed)
pdf("genetic_distance_histograms_allsites.pdf", width=12, height=10)  # Create a PDF file
par(mfrow=c(5,3), mar=c(4,4,3,1))  # 4 rows, 3 columns, adjust margins

# Plot each histogram with its title
hist(distgenDISS_AAR, main="AAR", xlab="Genetic Distance", col="lightblue")
hist(distgenDISS_ALD, main="Alderdice", xlab="Genetic Distance", col="lightblue")
hist(distgenDISS_BOU, main="Bouma", xlab="Genetic Distance", col="lightblue")
hist(distgenDISS_BRI, main="Bright", xlab="Genetic Distance", col="lightblue")
hist(distgenDISS_DIA, main="Diaphus", xlab="Genetic Distance", col="lightblue")
hist(distgenDISS_ELV, main="Elvers", xlab="Genetic Distance", col="lightblue")
hist(distgenDISS_FTR, main="FTR", xlab="Genetic Distance", col="lightblue")
hist(distgenDISS_GEY, main="Geyer", xlab="Genetic Distance", col="lightblue")
hist(distgenDISS_MCG, main="McGrail", xlab="Genetic Distance", col="lightblue")
hist(distgenDISS_RTR, main="RTR", xlab="Genetic Distance", col="lightblue")
hist(distgenDISS_EFGB, main="EFGB", xlab="Genetic Distance", col="lightblue")
hist(distgenDISS_WFGB, main="WFGB", xlab="Genetic Distance", col="lightblue")
hist(distgenDISS_EDI, main="Edisto", xlab="Genetic Distance", col="lightblue")
# Reset the plot parameters
#par(mfrow=c(1,1))

dev.off()

# Method 2: Using ggplot2 for more customization (recommended)
# Install required packages if not already installed
# install.packages(c("ggplot2", "gridExtra", "reshape2"))

library(ggplot2)
library(gridExtra)
library(reshape2)

# Create a function to convert distance matrix to dataframe for ggplot
matrix_to_df <- function(dist_matrix, site_name) {
  # Convert distance matrix to vector
  dist_vec <- as.vector(as.matrix(dist_matrix))
  # Remove diagonal elements (zeros) if it's a square matrix
  if(is.matrix(dist_matrix)) {
    dist_vec <- dist_vec[dist_vec > 0]
  }
  return(data.frame(site = site_name, distance = dist_vec))
}

# Combine all distance data into one dataframe
all_distances <- rbind(
  matrix_to_df(distgenDISS_AAR, "AAR"),
  matrix_to_df(distgenDISS_ALD, "Alderdice"),
  matrix_to_df(distgenDISS_BOU, "Bouma"),
  matrix_to_df(distgenDISS_BRI, "Bright"),
  matrix_to_df(distgenDISS_DIA, "Diaphus"),
  matrix_to_df(distgenDISS_EDI, "Edisto"),
  matrix_to_df(distgenDISS_EFGB, "EFGB"),
  matrix_to_df(distgenDISS_ELV, "Elvers"),
  matrix_to_df(distgenDISS_FTR, "FTR"),
  matrix_to_df(distgenDISS_GEY, "Geyer"),
  matrix_to_df(distgenDISS_MCG, "McGrail"),
  matrix_to_df(distgenDISS_RTR, "RTR"),
  matrix_to_df(distgenDISS_WFGB, "WFGB")
)

# Create the faceted plot
ggplot(all_distances, aes(x = distance)) +
  geom_histogram(fill = "lightblue", color = "black", bins = 30) +
  facet_wrap(~ site, scales = "free_y", ncol = 3) +
  labs(title = "Genetic Distance Distributions by Site",
       x = "Genetic Distance",
       y = "Count") +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "#3B8C6E"),
    strip.text = element_text(color = "white", face = "bold"),
    plot.title = element_text(hjust = 0.5, size = 16)
  )

# Save the plot
ggsave("genetic_distance_histograms_ggplot.pdf", width = 12, height = 10)
```


Convert SNP data into a genclone object
```{r SNPs to genclone object}
pc <- as.genclone(snps)
pc
```

#Perform clone filtering
Collapse multilocus genotypes (MLGs) based on genetic distance, using a threshold of 2000. 
The result reduces the 288 original genotypes to 189 multilocus genotypes
After further investigations into clone groups the clone corrected object will change
```{r}
mlg.filter(pc, threshold = 2000, missing = "asis", distance = "diss.dist")
mlg.filter(pc,  missing = "asis", distance = "diss.dist") <- 2000
pc
save(pc, file='./pc.Rdata')
```

# Look into initial identified clones
These clones will be further investigated and final clone samples will be defined through subsequent analyses
```{r}
# Get the original multilocus genotype assignments before filtering
original_mlg <- mlg.vector(snps)

# Get the current filtered multilocus genotype assignments
filtered_mlg <- mll(pc)

# Create a data frame to compare
comparison <- data.frame(
  sample_names = indNames(pc),
  original_mlg = original_mlg,
  filtered_mlg = filtered_mlg
)

# Find samples that were merged as clones
clone_groups_initial <- split(comparison$sample_names, comparison$filtered_mlg)

# Print groups where multiple samples were considered clones
for (group in clone_groups_initial) {
  if (length(group) > 1) {
    print("These samples were considered clones:")
    print(group)
  }
}

# Count original vs. filtered MLGs
print(paste("Original number of MLGs:", length(unique(original_mlg))))
print(paste("After filtering (at threshold 2000): 189 MLGs"))
```

# Look into clone group stats and summaries
```{r}
# 1. Initial setup and metadata preparation
se_metadata$sample_id <- rownames(se_metadata)

# Create initial clone IDs from your mlg.filter results
initial_clone.ids <- mlg.id(pc)

# 2. Create base clone analysis dataframe with initial clone IDs
clone_groups_initial <- data.frame(
  sample_id = rownames(se_metadata),
  se_metadata
) %>%
  left_join(
    data.frame(
      sample_id = unlist(initial_clone.ids),
      clone_group = rep(names(initial_clone.ids), sapply(initial_clone.ids, length))
    ),
    by = "sample_id"
  )

# 3. Create detailed clone summaries
clone_details_initial <- clone_groups_initial %>%
  group_by(clone_group) %>%
  filter(n() > 1) %>%
  arrange(clone_group, SITE) %>%
  summarise(
    samples = paste(sample_id, collapse=", "),
    sites = paste(unique(SITE), collapse=", "),
    depths = paste(sort(DEPTH), collapse=", ")
  )

# 4. Create comprehensive statistics
clone_stats_initial <- list(
  # Group level statistics
  group_summary = clone_groups_initial %>%
    group_by(clone_group) %>%
    summarize(
      n_samples = n(),
      n_sites = n_distinct(SITE),
      sites = paste(sort(unique(SITE)), collapse=", "),
      depth_range = paste(min(DEPTH), "-", max(DEPTH), "m"),
      sample_ids = paste(sample_id, collapse=", "),
      depths = paste(sort(DEPTH), collapse=", ")
    ) %>%
    filter(n_samples > 1),
  
  # Overall statistics
  overall_stats = list(
    total_groups = length(unique(clone_groups_initial$clone_group)),
    cloned_samples = sum(table(clone_groups_initial$clone_group) > 1),
    size_distribution = table(table(clone_groups_initial$clone_group)),
    original_samples = nInd(pc),
    corrected_samples = length(unique(clone_groups_initial$clone_group)),
    samples_per_pop = table(pop(pc))
  )
)

# 5. Save all results with "initial" in the filename
save(clone_groups_initial, clone_details_initial, clone_stats_initial, 
     file='initial_clone_analysis_results.Rdata')

# 6. Print summary information
print("Initial Clone Group Details:")
print(clone_details_initial, n=Inf)
print("\nInitial Clone Statistics:")
with(clone_stats_initial$overall_stats, {
  cat("Total initial clone groups:", total_groups, "\n")
  cat("Number of cloned samples in initial analysis:", cloned_samples, "\n")
  cat("Initial clone group size distribution:\n")
  print(size_distribution)
  cat("\nSamples before correction:", original_samples, "\n")
  cat("Unique multilocus genotypes after initial correction:", corrected_samples, "\n")
  cat("\nSamples per population in initial dataset:\n")
  print(samples_per_pop)
})
```



# Investigating clone groups with multiple sites
Clone group 182 included sites Geyer and EFGB
Clone group 221 included sites Geyer and EFGB
Clone group 195 included sites Elvers and Bouma
Examine whether samples that were clustered as clones but appear at different sites are truly genetically identical or if they might have been erroneously grouped due to the distance threshold (was originally 5000)
```{r}
# Create target_groups object for specific clone groups of interest (182, 2211, 195)
target_groups <- clone_groups_initial %>%
  filter(clone_group %in% c(182, 221, 195))

# Look at the structure of the distance matrix
dim(as.matrix(distgenDISS))
rownames(as.matrix(distgenDISS))[1:10]  # Check first few sample names

# Define matrix_samples as the rownames of the distance matrix
matrix_samples <- rownames(as.matrix(distgenDISS))

# Filter for only samples that exist in distance matrix
available_samples <- target_groups %>%
  filter(sample_id %in% matrix_samples)

# Available samples by clone group:
print(available_samples %>% 
      select(clone_group, sample_id, SITE, DEPTH) %>%
      arrange(clone_group))

# Get distances just for these samples
clone_distances <- as.matrix(distgenDISS)[available_samples$sample_id, available_samples$sample_id]
print("Genetic distances between available samples:")
print(clone_distances)

# Get context by comparing to overall distance distribution
dist_summary <- summary(as.vector(distgenDISS))
print("Overall distance distribution summary:")
print(dist_summary)
```

#Looking into pairs SH10553/SH10653 (distance=951) and SH10781/SH10817 (distance=1475)
```{r}
# Extract SNP data for these specific samples
pairs_to_check <- c("SH10553", "SH10653", "SH10781", "SH10817")

# Get the SNP data - first let's look at the format of our data
str(snps)

# Extract genotypes for our pairs
snp_data <- tab(snps)[pairs_to_check,]

# Compare SNPs between first pair
pair1_diff <- snp_data["SH10553",] != snp_data["SH10653",]
pair2_diff <- snp_data["SH10781",] != snp_data["SH10817",]

# Summary of differences
cat("Number of SNP differences between SH10553/SH10653:", sum(pair1_diff), "\n")
cat("Number of SNP differences between SH10781/SH10817:", sum(pair2_diff), "\n")

# Look at the actual differing positions
cat("First few SNP differences between SH10553/SH10653:\n")
head(which(pair1_diff))
print(snp_data[c("SH10553","SH10653"), which(pair1_diff)[1:10]])

cat("First few SNP differences between SH10781/SH10817:\n")
head(which(pair2_diff))
print(snp_data[c("SH10781","SH10817"), which(pair2_diff)[1:10]])
```

# Count SNP differences between the above pairs
```{r}
# Extract SNP data for these specific samples
pairs_to_check <- c("SH10553", "SH10653", "SH10781", "SH10817")
snp_data <- tab(snps)[pairs_to_check,]

# Function to count meaningful genotype differences
count_genotype_differences <- function(sample1, sample2) {
  # Convert the data to a format that makes comparison easier
  geno1 <- snp_data[sample1,]
  geno2 <- snp_data[sample2,]
  
  # Count differences - considering the actual genotype states
  differences <- sum(geno1 != geno2, na.rm=TRUE)
  
  # Get the locations of differences for examination
  diff_positions <- which(geno1 != geno2)
  
  return(list(
    total_differences = differences,
    diff_positions = diff_positions,
    sample_differences = cbind(
      geno1[diff_positions[1:min(10, length(diff_positions))]],
      geno2[diff_positions[1:min(10, length(diff_positions))]]
    )
  ))
}

# Compare both pairs
pair1_diffs <- count_genotype_differences("SH10553", "SH10653")
pair2_diffs <- count_genotype_differences("SH10781", "SH10817")

# Print results
cat("Differences between SH10553/SH10653:", pair1_diffs$total_differences, "\n")
cat("First few genotype differences:\n")
print(pair1_diffs$sample_differences)

cat("Differences between SH10781/SH10817:", pair2_diffs$total_differences, "\n")
cat("First few genotype differences:\n")
print(pair2_diffs$sample_differences)
```

# Checking SH10781 (Elvers) against all Bouma clones
```{r}
# Get all samples from clone group containing SH10781
clone_group_samples <- clone_groups_initial %>%
  filter(clone_group == clone_groups_initial$clone_group[clone_groups_initial$sample_id == "SH10781"]) %>%
  pull(sample_id)

print("All samples in this clone group:")
print(clone_group_samples)

# Get the data from original snps object
snp_data_full <- tab(snps)[clone_group_samples,]

# Modify our comparison function for the full dataset
count_genotype_differences <- function(sample1, sample2) {
  geno1 <- snp_data_full[sample1,]
  geno2 <- snp_data_full[sample2,]
  
  differences <- sum(geno1 != geno2, na.rm=TRUE)
  diff_positions <- which(geno1 != geno2)
  
  return(list(
    total_differences = differences,
    diff_positions = diff_positions,
    sample_differences = cbind(
      geno1[diff_positions[1:min(10, length(diff_positions))]],
      geno2[diff_positions[1:min(10, length(diff_positions))]]
    )
  ))
}

# Compare SH10781 to all other samples in the group
for(sample in clone_group_samples[clone_group_samples != "SH10781"]) {
  diffs <- count_genotype_differences("SH10781", sample)
  
  cat("\nDifferences between SH10781 and", sample, ":", diffs$total_differences, "\n")
  cat("First few genotype differences:\n")
  print(diffs$sample_differences[1:5,])
}
```

# Create final clone-corrected snps dataset
Includes individuals confirmed not to be true clones between sites
Retains only one individual per clone group to avoid bias from identical genotype
Removing Clone group 182 and 221 because their sample numbers are similar and likely got switched during sample processing/extraction. Removing Bouma sample from Elvers clone group 195 because it is a separate individual than the Elvers individuals in this group
```{r}
#library(poppr)
# Start with original data
pc <- as.genclone(snps)

# Set distance threshold for initial filtering
mlg.filter(pc, distance = "diss.dist") <- 2000

# Get current MLG assignments
current_mlgs <- mlg.vector(pc)
custom_mlgs <- current_mlgs

# Get the specific Elvers sample from group 195
elvers_195 <- clone_groups_initial %>%
  filter(clone_group == "195", SITE == "Elvers") %>%
  pull(sample_id)

# Get all samples from groups 221 and 182
samples_221_182 <- clone_groups_initial %>%
  filter(clone_group %in% c("221", "182")) %>%
  pull(sample_id)

# Combine all samples to separate
samples_to_separate <- c(elvers_195, samples_221_182)

# Give these samples unique MLGs
for(i in seq_along(samples_to_separate)) {
  custom_mlgs[which(indNames(pc) == samples_to_separate[i])] <- max(as.numeric(current_mlgs)) + i
}

# Set strata and perform clone correction
strata(pc) <- data.frame(pop = pop(pc))
pc.corr <- clonecorrect(pc, strata = ~pop)

# Convert to genind object
snps_mlg <- genclone2genind(pc.corr)

# Save the results
save(snps_mlg, file='./snps_mlg.Rdata')

# Verify the reduction in samples
print("Number of samples before correction:")
print(nInd(pc))
print("Number of MLGs before correction:")
print(length(unique(mlg.vector(pc))))
print("Number of samples after correction:")
print(nInd(pc.corr))
print("Number of MLGs after correction:")
print(length(unique(mlg.vector(pc.corr))))

# Verify the conversion
print(snps_mlg)
print("Samples per population:")
print(summary(snps_mlg$pop))
#NEW CLONE CORRECTED DATA
```


# Clone group mapping export & metadata correction
This sectionindependently re-detects clone groups per site (via pairwise genetic distance,
not poppr::mlg.filter) to build the objects used later for QGIS/mapping exports
and includes a necessary metadata correction:
 1. Helper functions for genetic distance analysis and network visualization
 2. Corrects a metadata coordinate error (SH11757 latitude)
 3. Site-by-site analysis identifying potential clones based on genetic similarity
 4. Consolidated dataframe linking samples to their clone groups (`clone_analysis_updated`)
 5. `clean_clone_details`: the clone catalog used later in the Diversity Table
    (built from `clone_details_initial`, the poppr-based clone catalog above —
    independent of the site-by-site redetection in this chunk)
 6. Saves results needed downstream

Object outputs used elsewhere in this script:
 final_clone_details, all_results, site_summary: feed the QGIS/mapping export
   chunks below ("Export Clone group information for QGIS", etc.)
 clone_analysis_updated: used in the mapping export chunks and the clone-distance
   scatterplot/histogram chunks below
 clean_clone_details: used in the Table 1 / Diversity Table `clone_groups_per_site`
   calculation, much further downstream
```{r}
#===== CLONE GROUP MAPPING EXPORT & METADATA CORRECTION =====

# 1. Setup and functions
# Helper function to prepare site data and visualization parameters
prepare_site_data <- function(site_name, threshold = 2000) {
  # Get site data
  site_snps <- snps[pop=site_name]
  n_samples <- length(indNames(site_snps))
  
  if(n_samples < 2) {
    return(list(valid = FALSE, message = paste("Skipping", site_name, "- not enough samples")))
  }
  
  # Calculate distances
  dist_mat <- as.matrix(diss.dist(site_snps, percent = FALSE, mat = FALSE))
  
  # Setup visualization parameters
  edge_colors <- matrix("grey80", n_samples, n_samples)
  edge_widths <- matrix(0.5, n_samples, n_samples)
  
  # Find close pairs
  close_pairs <- which(dist_mat < threshold & dist_mat > 0, arr.ind = TRUE)
  
  # Set colors and widths for close pairs if they exist
  if(length(close_pairs) > 0) {
    for(i in 1:nrow(close_pairs)) {
      edge_colors[close_pairs[i,1], close_pairs[i,2]] <- "darkgreen"
      edge_colors[close_pairs[i,2], close_pairs[i,1]] <- "darkgreen"
      edge_widths[close_pairs[i,1], close_pairs[i,2]] <- 2
      edge_widths[close_pairs[i,2], close_pairs[i,1]] <- 2
    }
  }
  
  return(list(
    valid = TRUE,
    site_snps = site_snps,
    dist_mat = dist_mat,
    n_samples = n_samples,
    close_pairs = close_pairs,
    edge_colors = edge_colors,
    edge_widths = edge_widths
  ))
}

# Function to create network plots
create_network_plot <- function(site_data, site_name, threshold = 2000, save_to_file = FALSE) {
  if(save_to_file) {
    pdf(file = paste0("clone_network_", site_name, ".pdf"))
  }
  
  qgraph(1/site_data$dist_mat, 
         layout='spring', 
         vsize=5, 
         minimum = 1/threshold,
         edge.color = site_data$edge_colors,
         edge.width = site_data$edge_widths,
         labels = rownames(site_data$dist_mat),
         title = paste("Clone Network -", site_name))
  
  if(save_to_file) {
    dev.off()
  }
}

# Main function to analyze site clones
analyze_site_clones <- function(site_name, threshold = 2000, save_plot = FALSE) {
  tryCatch({
    # Prepare site data
    site_data <- prepare_site_data(site_name, threshold)
    
    if(!site_data$valid) {
      cat("\n", site_data$message, "\n")
      return(NULL)
    }
    
    # Create network visualization
    cat("\nCreating network for", site_name, "\n")
    create_network_plot(site_data, site_name, threshold, save_plot)
    
    # Initialize clone groups
    clone_groups <- list()
    processed <- c()
    
    # Process close pairs if they exist
    if(length(site_data$close_pairs) > 0 && nrow(site_data$close_pairs) > 0) {
      for(i in 1:nrow(site_data$close_pairs)) {
        s1 <- rownames(site_data$dist_mat)[site_data$close_pairs[i,1]]
        s2 <- colnames(site_data$dist_mat)[site_data$close_pairs[i,2]]
        
        if(all(c(s1, s2) %in% processed)) next
        
        group_found <- FALSE
        for(j in seq_along(clone_groups)) {
          if(any(c(s1, s2) %in% clone_groups[[j]])) {
            clone_groups[[j]] <- unique(c(clone_groups[[j]], s1, s2))
            group_found <- TRUE
            break
          }
        }
        
        if(!group_found) {
          clone_groups[[length(clone_groups) + 1]] <- c(s1, s2)
        }
        
        processed <- unique(c(processed, s1, s2))
      }
    }
    
    # Create results dataframe
    if(length(clone_groups) > 0) {
      results <- data.frame(
        site = site_name,
        clone_group = paste(site_name, seq_along(clone_groups), sep="_"),
        samples = sapply(clone_groups, paste, collapse=", "),
        n_samples = sapply(clone_groups, length),
        stringsAsFactors = FALSE
      )
    } else {
      results <- data.frame(
        site = site_name,
        clone_group = character(0),
        samples = character(0),
        n_samples = integer(0),
        stringsAsFactors = FALSE
      )
    }
    
    # Print summary
    cat("\nResults for", site_name, ":\n")
    cat("Total samples:", nrow(site_data$dist_mat), "\n")
    cat("Number of clone groups:", length(clone_groups), "\n")
    
    if(length(clone_groups) > 0) {
      cat("Samples in clone groups:", length(unique(unlist(clone_groups))), "\n")
      cat("\nClone groups:\n")
      print(results)
      
      if(length(site_data$close_pairs) > 0) {
        cat("\nDistances between clone pairs:\n")
        print(summary(site_data$dist_mat[site_data$close_pairs]))
      }
    }
    
    return(results)
    
  }, error = function(e) {
    cat("\nError processing", site_name, ":", conditionMessage(e), "\n")
    return(NULL)
  })
}

# 2. Correct metadata errors
# Correct Diaphus sample SH11757 latitude
cat("Current value for SH11757 latitude:", se_metadata["SH11757", "LAT"], "\n")
se_metadata["SH11757", "LAT"] <- 28.08812
cat("New value for SH11757 latitude:", se_metadata["SH11757", "LAT"], "\n")

# 3. Initial site-by-site clone analysis
# Extract unique site names from the metadata
all_sites <- unique(se_metadata$SITE)
# Process all sites
all_results <- list()
cat("\nProcessing", length(all_sites), "sites...\n")

for(site in all_sites) {
  cat("\n=====================================")
  cat("\nAnalyzing", site)
  cat("\n=====================================\n")
  result <- analyze_site_clones(site)
  if(!is.null(result) && nrow(result) > 0) {
    all_results[[site]] <- result
  }
}

# Combine results
if(length(all_results) > 0) {
  final_clone_details <- do.call(rbind, all_results)
  
  # Print summary
  cat("\nFinal summary across all sites:\n")
  site_summary <- table(final_clone_details$site)
  print(site_summary)
  
  cat("\nTotal clone groups found:", nrow(final_clone_details), "\n")
  
  # Save results
  write.csv(final_clone_details, "all_sites_clone_analysis.csv", row.names = FALSE)
} else {
  cat("\nNo clone groups found in any site.\n")
}

# 4. Create consolidated clone analysis dataframe
# Create updated clone_analysis with correct coordinates
clone_analysis_updated <- data.frame(
  sample_id = rownames(se_metadata),
  se_metadata
) %>%
  left_join(
    data.frame(
      sample_id = unlist(strsplit(final_clone_details$samples, ", ")),
      clone_group = rep(final_clone_details$clone_group, 
                       sapply(strsplit(final_clone_details$samples, ", "), length))
    ),
    by = "sample_id"
  )

# Verify the update
cat("Dimensions of updated clone_analysis:", dim(clone_analysis_updated), "\n")
cat("Number of clone groups:", length(unique(na.omit(clone_analysis_updated$clone_group))), "\n")

# Check specific coordinates that were corrected
cat("\nVerifying corrected coordinates (e.g., SH11757):\n")
print(clone_analysis_updated[clone_analysis_updated$sample_id == "SH11757", 
                           c("sample_id", "LONG", "LAT", "DEPTH", "clone_group")])

# 5. Clean-up problematic clone groups
# `clean_clone_details` is the clone catalog actually used downstream (Diversity
# Table). It's built from `clone_details_initial` (the poppr-based catalog from
# the section above), not from `final_clone_details` computed in this chunk.
# First get the Bouma samples from groups 182 and 221
bouma_samples_182_221 <- clone_details_initial %>%
  filter(clone_group %in% c("182", "221")) %>%
  separate_rows(samples, sep = ", ") %>%
  separate_rows(sites, sep = ", ") %>%
  filter(sites == "Bouma") %>%
  group_by(clone_group) %>%
  summarise(
    samples = paste(samples, collapse = ", "),
    sites = "Bouma",
    depths = paste(sort(unique(depths)), collapse = ", ")
  )

# Now create our clean clone dataset
clean_clone_details <- clone_details_initial %>%
  filter(!clone_group %in% c("182", "221")) %>%  # Remove these groups initially
  filter(sites != "Bouma, Elvers") %>%  # Remove cross-site group
  bind_rows(bouma_samples_182_221)  # Add back just the Bouma samples

# 6. Save results needed downstream
save(final_clone_details, all_results, site_summary, 
     clone_analysis_updated, clean_clone_details, 
     file = "clone_analysis_final.RData")
```


# Scatterplot of mean distances within clone groups, with points colored by depth
```{r}
# First, prepare the data
library(tidyverse)
library(geosphere)  # For distm function
library(viridis)    # For viridis color scales

get_clone_distances <- function(clone_group, metadata) {
  samples <- unlist(strsplit(clone_group, ", "))
  
  coords <- metadata %>%
    filter(sample_id %in% samples) %>%
    select(LONG, LAT)
  
  if(nrow(coords) > 1) {
    dist_matrix <- distm(as.matrix(coords), fun = distHaversine)  # Using distm with Haversine
    distances <- dist_matrix[upper.tri(dist_matrix)]
    if(length(distances) > 0) {
      return(list(
        mean_distance = mean(distances),
        max_distance = max(distances),
        sample_depths = metadata$DEPTH[metadata$sample_id %in% samples]  # Changed from 'depths' to 'sample_depths'
      ))
    }
  }
  return(list(mean_distance = 0, max_distance = 0, sample_depths = metadata$DEPTH[metadata$sample_id %in% samples]))
}

# Process the data - exclude NA clone groups and composite sites
plot_data <- clean_clone_details %>%
  # Filter out rows with NA clone_group
  filter(!is.na(clone_group)) %>%
  # Also filter out any row where sites contains a comma
  filter(!grepl(",", sites)) %>%
  rowwise() %>%
  mutate(
    sample_list = list(unlist(strsplit(samples, ", "))),
    n_samples = length(sample_list),
    # Get depth information from clone_analysis_updated
    mean_depth = mean(clone_analysis_updated$DEPTH[clone_analysis_updated$sample_id %in% sample_list], na.rm = TRUE),
    # Calculate distances
    distances = list(get_clone_distances(samples, clone_analysis_updated))
  ) %>%
  unnest_wider(distances, names_sep = "_") %>%
  ungroup()

# Plot scatter plot
# Use the correct column names after unnest_wider
ggplot(plot_data, aes(x = reorder(sites, distances_mean_distance), y = distances_mean_distance, color = mean_depth)) + 
    geom_point(size = 5, alpha = 0.8) +
    scale_color_viridis(option = "mako", direction = -1, 
                      name = "Mean Depth (m)") +
    
    # Add the mean line across the entire plot
    geom_hline(yintercept = mean(plot_data$distances_mean_distance, na.rm = TRUE),
              color = "red", linetype = "dashed", size = 1.0) +
    # Add text label for the mean
    annotate("text", 
             x = 1, # Position at first site
             y = mean(plot_data$distances_mean_distance, na.rm = TRUE) + 1.5,
             label = paste("Mean =", round(mean(plot_data$distances_mean_distance, na.rm = TRUE), 2), "m"),
             color = "red", hjust = 0) +
    scale_y_continuous(
         limits = c(0, 30),  # Set limits to 0-30 meters
         breaks = seq(0, 30, 5),  # Breaks every 5 meters
         labels = function(x) format(x, digits = 1)
     ) +
     coord_flip() +  # Flips x and y axes
     labs(title = "Mean Clone Distances by Site",
          subtitle = "Points colored by mean depth of clone group",
          x = "",  
          y = "Mean Distance Between Clones (meters)") +
     theme_light() +
     theme(
         axis.text.y = element_text(size = 13, color = "black"),
         axis.text.x = element_text(size = 13, color = "black"),
         plot.title = element_text(size = 16, face = "bold"),
         plot.subtitle = element_text(size = 12, color = "gray40"),
         panel.grid.major.y = element_blank(),
         panel.grid.minor = element_blank(),
         legend.position = "right"
     )

# Save the plot
ggsave("clone_distances_depth.pdf", width = 9, height = 8)

# Print summary statistics
cat("\nSummary statistics by site (distances in meters):\n")
clone_spatial_summary_stats <- plot_data %>%
  group_by(sites) %>%
  summarise(
    n_clones = n(),
    mean_distance = round(mean(distances_mean_distance, na.rm = TRUE), 2),
    max_distance = round(max(distances_max_distance, na.rm = TRUE), 2),
    mean_depth = round(mean(mean_depth, na.rm = TRUE), 1)
  )
print(clone_spatial_summary_stats)
```

# Plot simple histogram of mean distances within clone groups
```{r}
pdf("clone_distances_histo.pdf", width = 9, height = 8) 
hist(plot_data$distances_mean_distance, 
     breaks = 10,  
     col = "gray",  
     border = "black",  
     xlab = "Mean Distance Within Clone Groups (m)", 
     main = ""  # No title
)
dev.off()
```

# Export Clone group information for QGIS
```{r}
# Create table with samples as rows
sample_table <- clone_analysis_updated %>%
  # Start with sample info
  select(sample_id, SITE, LONG, LAT, DEPTH) %>%
  # Join with clone group information
  left_join(
    final_clone_details %>%
      mutate(sample_id = strsplit(samples, ", ")) %>%
      unnest(sample_id) %>%
      select(clone_group, sample_id),
    by = "sample_id"
  ) %>%
  # Arrange by site and clone group
  arrange(SITE, clone_group, sample_id)

# Write as tab-delimited text file
write.table(sample_table, 
            "samples_with_clones.txt", 
            sep = "\t",            
            row.names = FALSE,     
            quote = FALSE)         

# Preview the table
head(sample_table)
```

# Table all singletons included in snps genid object before filtering
This can be used for mapping all sample points 
```{r}
# Print initial counts
cat("Initial samples in snps genind:", nInd(snps), "\n")
cat("Sample names in snps:", length(indNames(snps)), "\n")

# Create table with samples that are in snps
sample_table <- clone_analysis_updated %>%
  # Filter for samples that are in snps
  filter(sample_id %in% indNames(snps)) %>%
  # Start with sample info
  select(sample_id, SITE, LONG, LAT, DEPTH) %>%
  # Join with clone group information
  left_join(
    final_clone_details %>%
      mutate(sample_id = strsplit(samples, ", ")) %>%
      unnest(sample_id) %>%
      select(clone_group, sample_id),
    by = "sample_id"
  ) %>%
  # Arrange by site and clone group
  arrange(SITE, clone_group, sample_id)

# Print sample counts at different stages
cat("\nSamples in clone_analysis_updated:", nrow(clone_analysis_updated), "\n")
cat("Samples after filtering for snps:", nrow(sample_table), "\n")
cat("Samples with clone groups:", sum(!is.na(sample_table$clone_group)), "\n")

# Write as tab-delimited text file
write.table(sample_table, 
            "Swiftia_popgen_all_samples_postfilter.txt", 
            sep = "\t",            
            row.names = FALSE,     
            quote = FALSE)         

# Preview the table
head(sample_table)
dim(sample_table)
```

# Create table of just clone groups
With clone group categories 1-8 for mapping
```{r}
# Create table with numeric clone ID
clone_sample_table <- clone_analysis_updated %>%
  select(sample_id, SITE, LONG, LAT, DEPTH) %>%
  # Join with clone group information
  left_join(
    final_clone_details %>%
      mutate(sample_id = strsplit(samples, ", ")) %>%
      unnest(sample_id) %>%
      select(clone_group, sample_id),
    by = "sample_id"
  ) %>%
  # Remove rows where clone_group is NA
  filter(!is.na(clone_group)) %>%
  # Add new cloneID column by extracting number after underscore
  mutate(cloneID = sub(".*_", "", clone_group)) %>%
  # Arrange by site and clone group
  arrange(SITE, clone_group, sample_id)

# Write as tab-delimited text file
write.table(clone_sample_table, 
            "clone_samples_with_ID.txt", 
            sep = "\t",            
            row.names = FALSE,     
            quote = FALSE)         

# Preview the table
head(clone_sample_table)
```


# Checking Edisto distances to make sure we aren't missing any clones
```{r}
# Get Edisto samples
edisto_snps <- snps[pop='Edisto']

# Calculate distances
distgenDISS_ED <- diss.dist(edisto_snps, percent = FALSE, mat = FALSE)

# Create histogram
hist(distgenDISS_ED, 
     breaks = 30,
     main = "Distribution of Genetic Distances - Edisto",
     xlab = "Genetic Distance",
     ylab = "Frequency")

# Add vertical line at threshold
abline(v = 2000, col = "red", lwd = 2)
text(2000, max(hist(distgenDISS_ED, breaks = 30, plot = FALSE)$counts), 
     "Threshold (2000)", pos = 4, col = "red")

# Print summary statistics
cat("\nSummary statistics for Edisto distances:\n")
print(summary(distgenDISS_ED))

# Get number of pairs below threshold
n_below_threshold <- sum(distgenDISS_ED < 2000)
cat("\nNumber of pairs below threshold (2000):", n_below_threshold, "\n")

# Show closest pairs
if(n_below_threshold > 0) {
  close_pairs <- which(as.matrix(distgenDISS_ED) < 2000 & as.matrix(distgenDISS_ED) > 0, arr.ind = TRUE)
  if(nrow(close_pairs) > 0) {
    cat("\nPairs below threshold:\n")
    dist_matrix <- as.matrix(distgenDISS_ED)
    close_pairs_df <- data.frame(
      Sample1 = rownames(dist_matrix)[close_pairs[,1]],
      Sample2 = colnames(dist_matrix)[close_pairs[,2]],
      Distance = dist_matrix[close_pairs]
    )
    print(close_pairs_df[order(close_pairs_df$Distance),])
  }
}
```

#Euclidean distance b/w allele frequencies
```{r}
distgenEUCL <- dist(snps_mlg, method = "euclidean", diag = FALSE, upper = FALSE, p = 2)
hist(distgenEUCL)
save(distgenEUCL, file = './distgenEUCL_se.Rdata')
```

#Pairwise number of allelic differences between two individuals (diss.dist {poppr})
```{r}
distgenDISS <- diss.dist(snps_mlg, percent = FALSE, mat = FALSE)
hist(distgenDISS)
save(distgenDISS, file = './distgenDISS_se.Rdata')
#Visualize data via histograms and a network graph
qgraph(1/distgenDISS, layout='spring', vsize=5)
```

================================================================================
------------------------Further Data Filtering:He- HWequil.---------------------
================================================================================

#Convert genetic data to a hierfstat format and calcuate basic stats
```{r convert to hierfstat and basic stats}
# Check the distribution of Het
snps_mlg_hfs <- genind2hierfstat(snps_mlg) # Create hierfstat object

# Perform basic statistics
basicstat_all <- basic.stats(snps_mlg_hfs, diploid = TRUE, digits = 2)
save(basicstat_all, file='basicstat_all.Rdata')

# Extract heterozygosity data
Het_all <- as.data.frame(basicstat_all$perloc)

# Create histogram and check the distribution of observed heterozygosity (Ho)
ggplot(Het_all, aes(x=Ho))+
  geom_histogram(binwidth = 0.01,color="black", fill="white")+
  theme_classic()

# Check the markers with high heterozygosity Ho > 0.5
Het_tab_all <- Het_all[which(Het_all$Ho  > 0.5), ]
dim(Het_tab_all)
```

#Filter out markers with high Ho
```{r high Ho filtering}
#If any are found they can be removed by creating a vector of loci names and using adegenet to filter them out of dataset
toRemove_all <- which(Het_all$Ho  > 0.5)
snps_mlg_he = snps_mlg[loc=-toRemove_all]
snps_mlg_he
save(snps_mlg_he, file='snps_mlg_he.Rdata')
```

# Hardy-Weinberg Equilibrium (HWE)
Package: Pegas
Test for Hardy-Weinberg Equilibrium and remove loci that deviate significantly from HWE (p<0.01)
identifies loci that deviate from neutral expectations due to various factors including selection, but also non-selective forces like assortive mating, population structure, or genotyping errors
```{r HWE}
#library(pegas)
snps_mlg_he.hw <- hw.test(snps_mlg_he, B = 1000)
save(snps_mlg_he.hw, file='snps_mlg_he.hw.Rdata')
# View histogram of p-values
hist(snps_mlg_he.hw[,4], breaks=50, 
     main="Distribution of Hardy-Weinberg P-values",
     xlab="P-value", ylab="Frequency")
```

#Identify and remove loci not in HWE
```{r remove non-HWE loci}
loci_toRemove_hw <- which(snps_mlg_he.hw[,4] < 0.01)
length(loci_toRemove_hw)

# Create new neutral dataset
snps_mlg.neutral <- snps_mlg_he[loc=-loci_toRemove_hw]
snps_mlg.neutral
```

# Verify new HWE neutral dataset
```{r verify HWE neutral dataset}
# Check object types
class(snps_mlg_he)
class(snps_mlg.neutral)

# Verify genind object integrity
str(snps_mlg_he)

#Object's structure should confirm:
#Individuals: 192
#Loci: 23,605
#Populations: 13 (unique pop levels)
#Data Type:
#Codominant genetic data
#Diploid (ploidy = 2)

# Summarize p-value distribution
summary(snps_mlg_he.hw[,4])
```


===================================================
-----------------------BayeScan--------------------
===================================================
This section identifies loci under putative selection with BayeScan and
removes them to produce the neutral SNP panel used in all downstream
population-genetic analyses. Structure:
 1. Prepare data & run BayeScan (external program, run on the server)
 2. Load results and inspect basic diagnostics
 3. Apply the adopted outlier threshold (FST > 0.35, q < 0.05) and build the
    final neutral dataset
 4. Supplementary: alternative outlier-threshold explorations kept for
    documentation of the decision process (not used in the final pipeline)

# Prepare data for BayeScan
```{r prepare data}
#library(adegenet)
#library(dplyr)

genind2bayescan_efficient <- function(genind_obj, outfile) {
  # Get populations and loci info
  pops <- levels(genind_obj$pop)
  n_pops <- length(pops)
  n_loci <- nLoc(genind_obj)
  
  # Open file connection
  con <- file(outfile, "w")
  
  # Write header
  writeLines(paste("[loci]=", n_loci), con)
  writeLines(paste("[populations]=", n_pops), con)
  
  # Efficient processing
  for(loc in 1:n_loci) {
    # Get locus name
    loc_name <- locNames(genind_obj)[loc]
    
    # Extract allele data for this locus
    loc_data <- genind_obj@tab[, grep(paste0("^", loc_name), colnames(genind_obj@tab))]
    
    # Count total genes and allele frequencies
    total_genes <- nrow(loc_data) * 2
    allele_counts <- colSums(loc_data, na.rm = TRUE)
    
    # Ensure we have two allele counts
    if(length(allele_counts) >= 2) {
      # Write locus information
      writeLines(paste(loc, total_genes, "2", 
                       allele_counts[1], allele_counts[2]), 
                 con)
    }
  }
  
  # Close file connection
  close(con)
}

# Use the function
genind2bayescan_efficient(snps_mlg.neutral, "swiftia_snps_bayescan_neutral.txt")

# Verify file creation
file.info("swiftia_snps_bayescan_neutral.txt")
```

# Verify BayeScan file
```{r verify bscn file}
# Read first few lines
readLines("swiftia_snps_bayescan_neutral.txt", n=10)

# Count total lines
total_lines <- length(readLines("swiftia_snps_bayescan_neutral.txt"))

# Verify header and first few data lines
validate_bayescan_file <- function(filename) {
  lines <- readLines(filename)
  
  # Check header
  header_loci <- lines[1]
  header_pops <- lines[2]
  
  cat("Loci header:", header_loci, "\n")
  cat("Populations header:", header_pops, "\n")
  
  # Extract number of loci and populations
  n_loci <- as.numeric(gsub("\\[loci\\]=", "", header_loci))
  n_pops <- as.numeric(gsub("\\[populations\\]=", "", header_pops))
  
  cat("Number of loci:", n_loci, "\n")
  cat("Number of populations:", n_pops, "\n")
  cat("Total lines in file:", total_lines, "\n")
  
  # Check first few data lines
  data_lines <- lines[3:min(10, length(lines))]
  cat("\nFirst few data lines:\n")
  print(data_lines)
}

# Run validation
validate_bayescan_file("swiftia_snps_bayescan_neutral.txt")
```

# Generate population assignment file in R
```{r generate pop file}
# Create a numeric population assignment
pops <- levels(snps_mlg.neutral$pop)
pop_numeric <- as.numeric(factor(pop(snps_mlg.neutral), levels=pops))

# Write to file
write(pop_numeric, file = "swiftia_population_assignments.txt", ncolumns = 1)

# Verify contents
cat("First few lines of population assignments:\n")
head(readLines("swiftia_population_assignments.txt"))

# Verify total length and population distribution
length(pop_numeric)  # Should be 192
table(pop_numeric, pop(snps_mlg.neutral))
```

# If above population assignment does not work from R, use bash to create population assignment file:
```{bash alt create pop assignment file}
# Populations and their sizes
# Edisto: 16, AAR: 12, RTR: 26, Alderdice: 13, Geyer: 5, EFGB: 20, WFGB: 5, 
# Bouma: 13, Elvers: 18, Bright: 8, Diaphus: 22, McGrail: 16, FTR: 18

awk 'BEGIN {
    populations[1] = 16   # Edisto
    populations[2] = 12   # AAR
    populations[3] = 26   # RTR
    populations[4] = 13   # Alderdice
    populations[5] = 5    # Geyer
    populations[6] = 20   # EFGB
    populations[7] = 5    # WFGB
    populations[8] = 13   # Bouma
    populations[9] = 18   # Elvers
    populations[10] = 8   # Bright
    populations[11] = 22  # Diaphus
    populations[12] = 16  # McGrail
    populations[13] = 18  # FTR
}
{
    for(pop in populations) {
        for(i=1; i<=populations[pop]; i++) {
            print pop
        }
    }
}' /dev/null > swiftia_population_assignments_precise.txt
```


# Format the swiftia_snps_bayescan_neutral.txt file further for Bayescan
Creates a population-specific with complete locus coverage
This will take a few minutes on server
```{bash format txt file}
# Log into server (replace with your own username/host)
ssh <username>@<server_address>

# Transfer swiftia_snps_bayescan_neutral.txt to the server (e.g. via scp/rsync/SFTP client)
# Navigate to correct directory
cd /media/popgen/swiftia

# Format file
(
echo "[loci]=16764"
echo ""
echo "[populations]=13"
echo ""
pop_sizes=(16 12 26 13 5 20 5 13 18 8 22 16 18)
total_loci=16764
current_line=1

for pop in "${!pop_sizes[@]}"; do
    pop_num=$((pop+1))
    pop_size=${pop_sizes[$pop]}
    
    echo "[pop]=$pop_num"
    
    # Create an associative array to store locus data
    declare -A locus_data
    
    # First pass: read all data for this population into memory
    while IFS=' ' read -r locus_num _ _ allele1 allele2; do
        if [[ $locus_num =~ ^[0-9]+$ ]]; then
            locus_data[$locus_num]="$allele1 $allele2"
        fi
    done < swiftia_snps_bayescan_neutral.txt
    
    # Second pass: output data for each locus
    for ((locus=1; locus<=total_loci; locus++)); do
        # Progress indicator
        if ((locus % 1000 == 0)); then
            echo "Processing population $pop_num, locus $locus / $total_loci" >&2
        fi
        
        # Use the stored data if available, otherwise use zeros
        if [[ -n "${locus_data[$locus]}" ]]; then
            read allele1 allele2 <<< "${locus_data[$locus]}"
        else
            allele1=0
            allele2=0
        fi
        
        printf "  %d  384  2  %d  %d\n" $locus $allele1 $allele2
    done
    
    echo ""
    unset locus_data
done
) > swiftia_snps_bayescan_neutral_comprehensive.txt
```

# Run Bayescan
```{bash run bayescan}
#Find exact path to bayscan on server for following command
ls /opt/mambaforge/envs/popgen/bin/bayescan*

# Run using nohup, using 100 threads takes about 8 hours to run on server
nohup /opt/mambaforge/envs/popgen/bin/bayescan_2.1 swiftia_snps_bayescan_neutral_comprehensive.txt -population_prior swiftia_population_assignments_precise.txt -threads 100 -od . -o swiftia_bayescan_output

# Open additional terminal windo to server and view verify file
head -n 20 swiftia_bayescan_output_Verif.txt
```

Notes on BayeScan parameters used above:
- The burnin (-burn) and number of iterations (-n) might need adjustment based on your dataset
- The prior odds (-pr_odds) can be adjusted based on how many outliers you expect
- Monitor convergence using the chain files produced by BayeScan
- Consider running multiple times with different seeds to ensure consistency
- The q-value threshold (0.05) and FST threshold (0.35) below are the ones adopted for this manuscript; adjust based on how conservative you want to be

# Load BayeScan results
This is the canonical read of the BayeScan output — `results` is used
throughout the rest of this section, including the final outlier threshold
below.
```{r load bscn results}
library(ggplot2)

results <- read.table("swiftia_bayescan_output_fst.txt", header = TRUE)
colnames(results) <- c("prob", "log10.PO", "qval", "alpha", "fst")
results$locus <- 1:nrow(results)

# Diagnostic plots
ggplot(results, aes(x = fst, y = log10.PO)) +
  geom_point(alpha = 0.6) +
  theme_bw() +
  labs(x = "FST", y = "log10(PO)",
       title = "BayeScan Results: FST vs log10(PO)")

ggplot(results, aes(x = qval)) +
  geom_histogram(bins = 50) +
  geom_vline(xintercept = 0.05, color = "red", linetype = "dashed") +
  theme_bw() +
  labs(x = "q-value", y = "Count",
       title = "Distribution of q-values")

ggplot(results, aes(x = alpha)) +
  geom_histogram(bins = 50) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
  theme_bw() +
  labs(x = "Alpha", y = "Count",
       title = "Distribution of alpha values")

# Descriptive summary of the full results table (NOT an outlier count — the
# adopted outlier threshold is applied further below)
summary_stats <- data.frame(
  mean_fst = mean(results$fst),
  sd_fst = sd(results$fst),
  median_fst = median(results$fst),
  mean_alpha = mean(results$alpha),
  sd_alpha = sd(results$alpha),
  total_loci = nrow(results)
)
print(summary_stats)
write.csv(summary_stats, "bayescan_summary_stats.csv", row.names = FALSE)
```

# Plot highlighting the adopted outlier threshold
```{r plot outliers}
# Calculate -log10(q-value)
results$neg_log_qval <- -log10(results$qval)

# Create a new column to identify points in different regions
results$point_type <- case_when(
  results$fst > 0.35 ~ "high_outlier",  # Points in yellow box
  results$alpha > 0 ~ "diversifying",    # Other diversifying selection
  TRUE ~ "balancing"                     # Balancing selection
)

# Create the plot
plot <- ggplot(results, aes(x = neg_log_qval, y = fst)) +
  geom_point(aes(color = point_type), alpha = 0.6, size = 3) +
  scale_color_manual(values = c("blue", "red", "orange"),
                    labels = c("Balancing", "Diversifying", "High Outliers"),
                    name = "Selection Type") +
  theme_bw() +
  theme(panel.grid = element_blank()) +
  labs(x = "-log10(q-value)",
       y = "FST",
       title = "BayeScan Results") +
  geom_vline(xintercept = 1.3, linetype = "dashed")

# Save
ggsave("BayeScan_Results_Outliers.pdf", plot = plot, device = "pdf", 
       path = output_dir,
       scale = 1, width = 11, height = 10, units = "in", dpi = 300)
plot
```

# Apply final outlier threshold
High-confidence outliers under selection are removed at FST > 0.35 and q < 0.05.
This is the threshold adopted for the manuscript (expected: 18 outlier loci removed).
```{r appply final outlier threshold}
# NOTE: this vector was reconstructed to match the documented final pipeline
# threshold (FST > 0.35, q < 0.05, n = 18 outliers). Re-run and confirm
# length(outliers_to_remove) == 18 before finalizing.
outliers_to_remove <- which(results$fst > 0.35 & results$qval < 0.05)
length(outliers_to_remove)
```

# Outlier locus heatmap (adopted threshold)
Population-level allele frequencies at the loci actually removed
(`outliers_to_remove`, FST > 0.35 & q < 0.05), computed on `snps_mlg.neutral`
(the pre-outlier-removal object, so the outlier loci are still present).
```{r}
# Get the frequency table
freq_table <- tab(snps_mlg.neutral, freq = TRUE)

# Convert outlier indices to column pairs and filter out any that are too large
max_locus <- ncol(freq_table) / 2  # Maximum valid locus number
valid_outliers <- outliers_to_remove[outliers_to_remove <= max_locus]

# Convert valid outliers to column pairs
outlier_cols <- sort(c(valid_outliers * 2 - 1, valid_outliers * 2))

# Create matrix with outlier data
outlier_matrix <- freq_table[, outlier_cols]

# Verify dimensions
print("Number of valid outliers:")
print(length(valid_outliers))
print("Dimensions of outlier matrix:")
print(dim(outlier_matrix))

# Reshape to get one row per locus
n_outliers <- length(valid_outliers)
pop_freqs <- matrix(nrow = n_outliers, ncol = nPop(snps_mlg.neutral))
colnames(pop_freqs) <- levels(pop(snps_mlg.neutral))

# Calculate population means
for(i in 1:nPop(snps_mlg.neutral)) {
    pop_indices <- which(pop(snps_mlg.neutral) == levels(pop(snps_mlg.neutral))[i])
    # Take only first allele frequency for each locus
    pop_freqs[,i] <- colMeans(outlier_matrix[pop_indices, seq(1, ncol(outlier_matrix), 2)], na.rm=TRUE)
}

# Create heatmap
library(pheatmap)
pheatmap(pop_freqs,
         main = "Outlier Loci Allele Frequencies Across Populations (FST > 0.35, q < 0.05)",
         show_rownames = FALSE,
         cluster_rows = TRUE,
         cluster_cols = TRUE)

# Calculate and print summary statistics
print("Mean outlier allele frequencies per population:")
print(colMeans(pop_freqs))

outlier_var <- apply(pop_freqs, 1, var)
print("Top most variable outlier loci:")
print(head(sort(outlier_var, decreasing=TRUE), 10))

# Save summary
outlier_summary <- data.frame(
    locus_id = valid_outliers,
    mean_freq = rowMeans(pop_freqs),
    var_freq = outlier_var
)
write.csv(outlier_summary, "outlier_summary.csv", row.names=FALSE)
```

# Create final neutral dataset
```{r final neutral dataset}
# Create final neutral dataset
neutral_dataset_final <- snps_mlg.neutral[loc=-outliers_to_remove]
save(neutral_dataset_final, file="neutral_dataset_final.Rdata")

# Create neutral dataset metadata
# Extract individual names from the genind object
genind_samples_neu <- rownames(neutral_dataset_final@tab)

# Subset se_metadata to include only matching samples
se_metadata_neutral_all <- se_metadata[rownames(se_metadata) %in% genind_samples_neu, ]

# Create population map for downstream analyses
# Define all populations
pop_order_all <- c("Edisto", "AAR", "RTR", "Alderdice", "Geyer", "EFGB", "WFGB", "Bouma", "Elvers", 
               "Bright", "Diaphus", "McGrail", "FTR")

# Get the population for each sample
all_pops <- as.character(neutral_dataset_final@pop)

# Create subset_samples_all by selecting samples belonging to all populations
subset_samples_all <- genind_samples_neu[all_pops %in% pop_order_all]

# Check 
length(subset_samples_all)

# Create pop_code_all using subset_samples_all
pop_code_all <- data.frame(
  SAMPLE = subset_samples_all,
  STRATA = as.character(neutral_dataset_final@pop[match(subset_samples_all, rownames(neutral_dataset_final@tab))]),
  stringsAsFactors = FALSE
)

# Set row names for easier subsetting
rownames(pop_code_all) <- pop_code_all$SAMPLE

# Check the result
head(pop_code_all) 
#                     SAMPLE STRATA
# 14-v-18-1-001 14-v-18-1-001 Edisto
# 14-v-18-1-002 14-v-18-1-002 Edisto
# 14-v-18-1-004 14-v-18-1-004 Edisto
dim(pop_code_all)

# Create west to east order 
WE_order_all <- c("WFGB", "EFGB", "Bright", "Geyer", "Elvers", "McGrail", "Bouma", "Alderdice", "Diaphus", "FTR", "AAR", "RTR", "Edisto")
```


# -----------Subset data for downstream analyses------------------
```{r create data subsets}
# Subset with Edisto and all GoM populations with > 10 individuals (_sub)
snps_mlg_sub_n <- popsub(neutral_dataset_final, exclude=c('WFGB','Bright','Geyer'))
# Subset metadata (_sub)
se_metadata_neutral_sub <- subset(se_metadata_neutral_all, 
                                    !(SITE %in% c("WFGB", "Bright", "Geyer")))
save(se_metadata_neutral_sub, file='se_metadata_neutral_sub.Rdata')


# Subset without Edisto, all GoM populations (_GoMa)
snps_mlg_gom_all_n <- popsub(neutral_dataset_final, exclude = c('Edisto'))
save(snps_mlg_gom_all_n, file = 'snps_mlg_gom_all_n.Rdata')
# Subset metadata (_GoMa)
se_metadata_neutral_GoMa <- subset(se_metadata_neutral_all, SITE != "Edisto")
save(se_metadata_neutral_GoMa, file='se_metadata_neutral_GoMa.Rdata')


# Subset without Edisto and populations with less than 10 individuals (_GoM10)
snps_mlg_GoM10_n <- popsub(neutral_dataset_final, exclude=c('WFGB','Bright','Geyer','Edisto'))
save(snps_mlg_GoM10_n, file='snps_mlg_GoM10_n.Rdata')
# Subset metadata (_GoM10)
se_metadata_neutral_GoM10 <- subset(se_metadata_neutral_all, 
                                    !(SITE %in% c("WFGB", "Bright", "Geyer", "Edisto")))
save(se_metadata_neutral_GoM10, file='se_metadata_neutral_GoM10.Rdata')


# Create population maps for downstream analyses
# Function to create pop_code objects from a genind object
create_pop_code <- function(genind_obj, output_name) {
  # Extract sample names from the genind object
  samples <- rownames(genind_obj@tab)
  
  # Get the population for each sample
  pops <- as.character(genind_obj@pop)
  
  # Create the pop_code data frame
  pop_code <- data.frame(
    SAMPLE = samples,
    STRATA = pops,
    stringsAsFactors = FALSE
  )
  
  # Set row names for easier subsetting
  rownames(pop_code) <- pop_code$SAMPLE
  
  # Print summary statistics
  cat(paste("\nCreated", output_name, "with", nrow(pop_code), "samples\n"))
  cat("Population breakdown:\n")
  print(table(pop_code$STRATA))
  
  # Return the pop_code object
  return(pop_code)
}

# 1 For snps_mlg_sub_n (Subset with Edisto and all GoM populations with > 10 individuals)
pop_code_sub <- create_pop_code(snps_mlg_sub_n, "pop_code_sub")
save(pop_code_sub, file="pop_code_sub.Rdata")
pop_order_sub <- c("Edisto", "AAR", "RTR", "Alderdice", "EFGB", "Bouma", "Elvers", "Diaphus", "McGrail", "FTR")
WE_order_sub <- c("EFGB", "Elvers", "McGrail", "Bouma", "Alderdice", "Diaphus", "FTR", "AAR", "RTR", "Edisto")

# 2 For snps_mlg_gom_all_n (Subset without Edisto, all GoM populations)
pop_code_GoMa <- create_pop_code(snps_mlg_gom_all_n, "pop_code_GoMa")
save(pop_code_GoMa, file="pop_code_GoMa.Rdata")
pop_order_GoMa <- c("AAR", "RTR", "Alderdice", "Geyer", "EFGB", "WFGB", "Bouma", "Elvers", "Bright", "Diaphus", "McGrail", "FTR")
WE_order_GoMa <- c("WFGB", "EFGB", "Bright", "Geyer", "Elvers", "McGrail", "Bouma", "Alderdice", "Diaphus", "FTR", "AAR", "RTR")

# 3 For snps_mlg_GoM10 (Subset without Edisto and populations with less than 10 individuals)
pop_code_GoM10 <- create_pop_code(snps_mlg_GoM10_n, "pop_code_GoM10")
save(pop_code_GoM10, file="pop_code_GoM10.Rdata")
pop_order_GoM10 <- c("AAR", "RTR", "Alderdice", "EFGB", "Bouma", "Elvers", "Diaphus", "McGrail", "FTR")
WE_order_GoM10 <- c("EFGB", "Elvers", "McGrail", "Bouma", "Alderdice", "Diaphus", "FTR", "AAR", "RTR")

# Print comparison of sample sizes
cat("\nComparison of sample sizes:\n")
cat("pop_code_sub: ", nrow(pop_code_sub), "samples\n")
cat("pop_code_GoMa: ", nrow(pop_code_GoMa), "samples\n")
cat("pop_code_GoM10: ", nrow(pop_code_GoM10), "samples\n")

#Comparison of sample sizes:
#pop_code_sub:  174 samples
#pop_code_GoMa:  176 samples
#pop_code_GoM10:  158 samples

# Save all objects into a single .Rdata file
save(
  # Dataset objects
  snps_mlg_sub_n,
  snps_mlg_gom_all_n,
  snps_mlg_GoM10_n,
  
  # Metadata objects
  se_metadata_neutral_sub,
  se_metadata_neutral_GoMa,
  se_metadata_neutral_GoM10,
  
  # Population code objects
  pop_code_sub,
  pop_code_GoMa,
  pop_code_GoM10,
  
  # Population order vectors
  pop_order_sub,
  WE_order_sub,
  pop_order_GoMa,
  WE_order_GoMa,
  pop_order_GoM10,
  WE_order_GoM10,
  
  # Function (optional, but can be useful)
  create_pop_code,
  
  file = "nonclone_subsetted_data.Rdata"
)
```

#-------------------Summary of subsetted objects------------------
All populations (_all) 
192 Individuals
genid object: neutral_dataset_final
population mapping: pop_code_all
metadata: se_metadata_neutral_all
Pop order: pop_order_all ("Edisto", "AAR", "RTR", "Alderdice", "Geyer", "EFGB", "WFGB", "Bouma", "Elvers", 
               "Bright", "Diaphus", "McGrail", "FTR")
W2E order object: WE_order_all ("WFGB", "EFGB", "Bright", "Geyer", "Elvers", "McGrail", "Bouma", "Alderdice", "Diaphus", "FTR", "AAR", "RTR", "Edisto")

Edisto and GoM populations > 10 individuals (_sub)
174 samples
genid object: snps_mlg_sub_n 
population map: pop_code_sub
metadata: se_metadata_neutral_sub
pop order: pop_order_sub ("Edisto", "AAR", "RTR", "Alderdice", "EFGB", "Bouma", "Elvers", "Diaphus", "McGrail", "FTR")
W2E order object: WE_order_sub 

All GoM samples (_GoMa)
176 samples
genid object: snps_mlg_gom_all_n 
population map: pop_code_GoMa
metadata: se_metadata_neutral_GoMa
pop order: pop_order_GoMa ("AAR", "RTR", "Alderdice", "Geyer", "EFGB", "WFGB", "Bouma", "Elvers", "Bright", "Diaphus", "McGrail", "FTR")
W2E order object: WE_order_GoMa ("WFGB", "EFGB", "Bright", "Geyer", "Elvers", "McGrail", "Bouma", "Alderdice", "Diaphus", "FTR", "AAR", "RTR")

GoM populations greater than 10 individuals (_GoM10)
158 samples
snps_mlg_GoM10_n → pop_code_GoM10
se_metadata_neutral_GoM10
pop order: pop_order_GoM10 ("AAR", "RTR", "Alderdice" "EFGB", "Bouma", "Elvers", "Diaphus", "McGrail", "FTR")
W2E order object: WE_order_GoM10 ("EFGB", "Elvers", "McGrail", "Bouma", "Alderdice", "Diaphus", "FTR", "AAR", "RTR")

# Table 1: Filtering table
Computed directly from the pipeline objects created above, so it stays in sync
if any upstream filtering step changes. The LD-pruned (ADMIXTURE-only) rows are
recorded manually since LD pruning is done externally in PLINK, not in R.
```{r filtering table}
library(dplyr)

# Helper to summarize a genind object at each filtering step
summarize_step <- function(step, description, obj = NULL, n_ind = NA, n_loc = NA, n_pop = NA) {
  if (!is.null(obj)) {
    n_ind <- tryCatch(nInd(obj), error = function(e) NA)
    n_loc <- tryCatch(nLoc(obj), error = function(e) NA)
    n_pop <- tryCatch(length(unique(pop(obj))), error = function(e) NA)
  }
  data.frame(
    Step = step,
    Description = description,
    Individuals = n_ind,
    Loci = n_loc,
    Populations = n_pop,
    stringsAsFactors = FALSE
  )
}

filtering_table <- bind_rows(
  summarize_step("Raw dataset (post-import)",
                  "SNPs and individuals retained after removing loci with >10% missing data and individuals with >30% missing data",
                  obj = snps),

  summarize_step("Clone correction",
                  "One representative individual retained per confirmed clone group (clonecorrect); anomalous cross-site clone groups 182, 221, and one member of 195 removed",
                  obj = snps_mlg),

  summarize_step("Ho filtering",
                  paste0("Loci with observed heterozygosity Ho > 0.5 removed (n = ", length(toRemove_all), ")"),
                  obj = snps_mlg_he),

  summarize_step("HWE filtering",
                  paste0("Loci deviating from Hardy-Weinberg Equilibrium at p < 0.01 removed (hw.test, 1000 permutations); n removed = ", length(loci_toRemove_hw)),
                  obj = snps_mlg.neutral),

  summarize_step("Outlier removal (BayeScan)",
                  paste0("High-confidence outlier loci under selection removed (FST > 0.35, q < 0.05); n outliers = ", length(outliers_to_remove)),
                  obj = neutral_dataset_final),

  summarize_step("Final neutral dataset",
                  "Clone-corrected, neutral dataset used for downstream analyses",
                  obj = neutral_dataset_final),

  summarize_step("GoM10 subset", "Gulf populations n >= 10; Edisto excluded", obj = snps_mlg_GoM10_n),
  summarize_step("GoM10 (LD-pruned, ADMIXTURE only)", "PLINK --indep-pairwise 50 5 0.1", n_ind = 158, n_loc = 5419, n_pop = 9),

  summarize_step("Sub (Gulf-Atlantic) subset", "Ten populations incl. Atlantic outgroup (Edisto)", obj = snps_mlg_sub_n),
  summarize_step("Sub (LD-pruned, ADMIXTURE only)", "PLINK --indep-pairwise 50 5 0.1", n_ind = 174, n_loc = 5629, n_pop = 10),

  summarize_step("GoMa subset", "All Gulf populations regardless of size; Edisto excluded", obj = snps_mlg_gom_all_n),
  summarize_step("GoMa (LD-pruned, ADMIXTURE only)", "PLINK --indep-pairwise 50 5 0.1", n_ind = 176, n_loc = 5634, n_pop = 12)
)

print(filtering_table)
write.csv(filtering_table, "filtering_table_FINAL.csv", row.names = FALSE)
```

# Diversity Table
```{r diversity table}
library(hierfstat); library(poppr); library(dplyr)

# Verified 192-individual objects, loaded fresh
e192 <- new.env(); load("snps_mlg.Rdata", envir = e192)
stopifnot(nInd(e192$snps_mlg) == 192)
snps_mlg_192 <- e192$snps_mlg
stopifnot(nInd(neutral_dataset_final) == 192)  # your final analysis object

# N raw (pre-correction, from `snps`, 288-based)
n_raw <- table(pop(snps))

# N clone-corrected = N analyzed now (both trace to the same 192-lineage — no more mismatch)
n_clone_corrected <- table(pop(snps_mlg_192))
n_analyzed <- table(pop(neutral_dataset_final))

# Clone groups per site — using ONLY the 28 validated groups (row 29 excluded)
clone_groups_per_site <- clean_clone_details %>%
  filter(!is.na(clone_group)) %>%
  distinct(clone_group, sites) %>%
  count(sites, name = "n_clone_groups") %>%
  rename(SITE = sites)

# Coordinates / depth
site_geo <- se_metadata %>%
  group_by(SITE) %>%
  summarise(lat_mean = mean(LAT, na.rm=TRUE), long_mean = mean(LONG, na.rm=TRUE),
            depth_min = min(DEPTH, na.rm=TRUE), depth_max = max(DEPTH, na.rm=TRUE))

# Diversity stats on the CORRECT 192-individual final object
hfs_final <- genind2hierfstat(neutral_dataset_final)
basicstat_final <- basic.stats(hfs_final, diploid = TRUE, digits = 4)
ho_per_pop <- colMeans(basicstat_final$Ho, na.rm = TRUE)
he_per_pop <- colMeans(basicstat_final$Hs, na.rm = TRUE)
fis_per_pop <- colMeans(basicstat_final$Fis, na.rm = TRUE)
ar_per_pop <- colMeans(allelic.richness(hfs_final)$Ar, na.rm = TRUE)

# FIX 1: % polymorphic — explicitly verify subsetting is per-site, not whole-dataset
poly_per_pop <- sapply(levels(pop(neutral_dataset_final)), function(p) {
  sub <- neutral_dataset_final[pop = p]
  stopifnot(nInd(sub) == sum(pop(neutral_dataset_final) == p))  # guard
  mean(isPoly(sub, by = "locus", thres = 0)) * 100
})
print(poly_per_pop)  # sanity check — should now show real variation, not ~100% everywhere

# FIX 2: private alleles — verify grouping column before joining
priv <- poppr::private_alleles(neutral_dataset_final, report = "data.frame")
str(priv)  # confirm "population" column matches SITE labels exactly before joining
priv_per_pop <- priv %>% group_by(population) %>% summarise(n_private = n_distinct(allele))
print(priv_per_pop)  # sanity check — values should differ across sites now

# Assemble
site_diversity_table <- site_geo %>%
  left_join(as.data.frame(n_raw) %>% rename(SITE=Var1, N_raw=Freq), by="SITE") %>%
  left_join(as.data.frame(n_clone_corrected) %>% rename(SITE=Var1, N_clone_corrected=Freq), by="SITE") %>%
  left_join(as.data.frame(n_analyzed) %>% rename(SITE=Var1, N_analyzed=Freq), by="SITE") %>%
  left_join(clone_groups_per_site, by="SITE") %>%
  mutate(Ho=ho_per_pop[SITE], He=he_per_pop[SITE], FIS=fis_per_pop[SITE],
         Allelic_richness=ar_per_pop[SITE], Pct_polymorphic=poly_per_pop[SITE]) %>%
  left_join(priv_per_pop %>% rename(SITE=population), by="SITE")

print(site_diversity_table, n = Inf)
write.csv(site_diversity_table, "site_diversity_table_v2.csv", row.names = FALSE)
```

# Write out metadata files for all datasubsets
```{r export csv files}
# Master metadata (matches the raw/all-SNPs genind objects: snps, genind_all)
write.csv(se_metadata, file = "se_metadata_all.csv", row.names = TRUE)

# Neutral dataset metadata subsets (matches neutral_dataset_final and its population subsets)
write.csv(se_metadata_neutral_all,   file = "se_metadata_neutral_all.csv",   row.names = TRUE)
write.csv(se_metadata_neutral_sub,   file = "se_metadata_neutral_sub.csv",   row.names = TRUE)
write.csv(se_metadata_neutral_GoM10, file = "se_metadata_neutral_GoM10.csv", row.names = TRUE)
write.csv(se_metadata_neutral_GoMa,  file = "se_metadata_neutral_GoMa.csv",  row.names = TRUE)

# Trim metadata to match the 311 samples actually in the raw genind/VCF
se_metadata_all_snps <- se_metadata[rownames(se_metadata) %in% indNames(genind_all), ]
nrow(se_metadata_all_snps)  # should be 311

write.csv(se_metadata_all_snps, file = "se_metadata_all_snps.csv", row.names = TRUE)
```

